# Bayesian Econometrics {#sec-bayes}

::: {.callout-note}
## Chapter Status
This chapter is currently a placeholder for future Bayesian econometrics content.

Planned topics:
- Bayesian inference fundamentals
- Prior specification
- Posterior distributions
- MCMC methods
- Bayesian regression with `brms`
:::

```{r}
#| label: setup-bayes
#| include: false
# Packages: MCMCpack, brms available in before-chapter.R
```

## Introduction

This section serves as an introduction to Bayesian statistics and econometrics.

::: {.callout-tip}
## Recommended Resources
For Bayesian econometrics in R:

- [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/) by Richard McElreath
- [brms package](https://paul-buerkner.github.io/brms/) documentation
- [Bayesian Data Analysis](http://www.stat.columbia.edu/~gelman/book/) by Gelman et al.
:::

## Bayesian vs Frequentist

::: {.panel-tabset}

## Frequentist Approach

- Parameters are **fixed** but unknown
- Probability describes long-run frequency
- Inference via sampling distributions
- Confidence intervals

## Bayesian Approach

- Parameters are **random** variables
- Probability describes degree of belief
- Inference via Bayes' theorem
- Credible intervals
:::

## Bayes' Theorem

$$
p(\theta | y) = \frac{p(y | \theta) p(\theta)}{p(y)}
$$ {#eq-bayes-theorem}

Where:

- $p(\theta | y)$ is the **posterior** distribution
- $p(y | \theta)$ is the **likelihood**
- $p(\theta)$ is the **prior** distribution
- $p(y)$ is the marginal likelihood (normalizing constant)

::: {.callout-important}
## Key Insight
The posterior combines prior beliefs with data evidence:

**Posterior ∝ Likelihood × Prior**
:::

## Bayesian Linear Regression

Simple linear regression in Bayesian framework:

$$
\begin{aligned}
y_i &\sim N(\beta_0 + \beta_1 x_i, \sigma^2) \\
\beta_0, \beta_1 &\sim N(0, 10^2) \\
\sigma &\sim \text{Half-Cauchy}(0, 5)
\end{aligned}
$$ {#eq-bayes-lm}

```{r}
#| label: bayes-lm-example
#| eval: false
library(brms)

# Bayesian linear regression
bayes_model <- brm(
  y ~ x,
  data = mydata,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "b"),
    prior(cauchy(0, 5), class = "sigma")
  ),
  chains = 4,
  iter = 2000
)

summary(bayes_model)
```

## Further Development

This chapter will be expanded to include:

- [ ] Prior selection and sensitivity analysis
- [ ] Posterior predictive checks
- [ ] Bayesian model comparison (Bayes factors, LOO-CV)
- [ ] Hierarchical/multilevel models
- [ ] Practical applications in econometrics

::: {.callout-note}
## Contribute!
If you're interested in expanding this chapter, please see [CONTRIBUTING.md](CONTRIBUTING.md).
:::
