# Qualitative and Limited Dependent Variable Models {#sec-glm}

::: {.callout-note}
## Chapter Overview
This chapter covers models for qualitative and limited dependent variables:

- Binary choice models (Logit, Probit)
- Multinomial and conditional logit
- Ordered choice models
- Count data models (Poisson, Negative Binomial)
- Censored and truncated models (Tobit, Heckman)
:::

```{r}
#| label: setup-glm
#| include: false
# Packages loaded in before-chapter.R
# GLM-specific: nnet, survival, sampleSelection, censReg
```

## Generalized Linear Models Framework

Binary dependent variables are frequently studied in applied economics. Because a dummy variable $y$ can only take values 0 and 1, its (conditional) expected value equals the (conditional) probability that $y=1$:

$$
E(y|x) = 0 \times P(y = 0|x) + 1 \times P(y = 1|x) = P(y=1|x)
$$ {#eq-binary-expectation}

An important class of models specifies the success probability as:

$$
P(y = 1 | x) = G(\beta_0 + \beta_1 x_1 + \dots + \beta_k x_k) = G(\mathbf{x}'\boldsymbol{\beta})
$$ {#eq-glm-general}

::: {.callout-important}
## Link Functions
GLMs relate the mean of the response to a linear combination of predictors through a **link function** $g()$:

$$
\eta(\mathbf{x}) = g\left(\text{E}[Y \mid \mathbf{X} = \mathbf{x}]\right)
$$

The mean is then: $\text{E}[Y \mid \mathbf{X} = \mathbf{x}] = g^{-1}(\eta(\mathbf{x}))$
:::

### Common GLM Families {#sec-glm-families}

::: {.panel-tabset}

## Linear Regression

- **Distribution**: Normal $N(\mu(\mathbf{x}), \sigma^2)$
- **Link**: Identity $\eta(\mathbf{x}) = \mu(\mathbf{x})$
- **Use**: Numeric continuous data

## Logistic Regression

- **Distribution**: Bernoulli $\text{Bern}(p(\mathbf{x}))$
- **Link**: Logit $\eta(\mathbf{x}) = \log\left(\frac{p(\mathbf{x})}{1-p(\mathbf{x})}\right)$
- **Mean**: $p(\mathbf{x}) = \frac{1}{1 + e^{-\eta(\mathbf{x})}}$
- **Use**: Binary outcomes

## Probit Regression

- **Distribution**: Bernoulli $\text{Bern}(p(\mathbf{x}))$
- **Link**: Probit (inverse normal CDF)
- **Mean**: $p(\mathbf{x}) = \Phi(\eta(\mathbf{x}))$
- **Use**: Binary outcomes (normal latent variable)

## Poisson Regression

- **Distribution**: Poisson $\text{Pois}(\lambda(\mathbf{x}))$
- **Link**: Log $\eta(\mathbf{x}) = \log(\lambda(\mathbf{x}))$
- **Mean**: $\lambda(\mathbf{x}) = e^{\eta(\mathbf{x})}$
- **Use**: Count data

:::

## Binary Choice Models {#sec-binary-choice}

### Linear Probability Model (LPM)

The simplest approach uses OLS with a binary dependent variable:

$$
y_i = \mathbf{x}_i'\boldsymbol{\beta} + \epsilon_i
$$ {#eq-lpm}

::: {.callout-warning}
## LPM Limitations
1. Predicted probabilities can be < 0 or > 1
2. Heteroskedastic errors by construction
3. Linear marginal effects (unrealistic)

**Use robust standard errors!**
:::

### Logit Model

The logit model uses the logistic CDF:

$$
P(y=1|\mathbf{x}) = \frac{\exp(\mathbf{x}'\boldsymbol{\beta})}{1 + \exp(\mathbf{x}'\boldsymbol{\beta})} = \frac{1}{1 + \exp(-\mathbf{x}'\boldsymbol{\beta})}
$$ {#eq-logit}

```{r}
#| label: logit-example
#| eval: false
# Logit model
data("mroz", package = "wooldridge")
logit_model <- glm(inlf ~ nwifeinc + educ + exper + I(exper^2) +
                     age + kidslt6 + kidsge6,
                   family = binomial(link = "logit"), data = mroz)
summary(logit_model)
```

### Probit Model

The probit model uses the standard normal CDF:

$$
P(y=1|\mathbf{x}) = \Phi(\mathbf{x}'\boldsymbol{\beta})
$$ {#eq-probit}

```{r}
#| label: probit-example
#| eval: false
# Probit model
probit_model <- glm(inlf ~ nwifeinc + educ + exper + I(exper^2) +
                      age + kidslt6 + kidsge6,
                    family = binomial(link = "probit"), data = mroz)
summary(probit_model)
```

::: {.callout-tip}
## Logit vs Probit
- Very similar predicted probabilities in practice
- Logit has fatter tails (more weight on extreme values)
- Probit assumes normal latent variable
- **Convention**: Use logit unless theory suggests probit
:::

### Marginal Effects {#sec-marginal-effects}

Coefficients in logit/probit are not directly interpretable. We compute **marginal effects**:

$$
\frac{\partial P(y=1|\mathbf{x})}{\partial x_k} = g(\mathbf{x}'\boldsymbol{\beta}) \beta_k
$$ {#eq-marginal-effects}

where $g()$ is the PDF of the link function.

::: {.panel-tabset}

## At Means (MEM)

Evaluate at mean values of $\mathbf{x}$:

```{r}
#| label: mem-logit
#| eval: false
library(margins)
margins(logit_model, at = list(x = mean(x)))
```

## Average Marginal Effects (AME)

Average marginal effects across all observations:

```{r}
#| label: ame-logit
#| eval: false
margins(logit_model)  # AME is the default
```

:::

::: {.callout-note}
**Best Practice**: Report Average Marginal Effects (AME) for better interpretability.
:::

## Multinomial Models {#sec-multinomial}

### Multinomial Logit

For unordered choices with $J > 2$ alternatives:

$$
P(y_i = j | \mathbf{x}_i) = \frac{\exp(\mathbf{x}_i'\boldsymbol{\beta}_j)}{\sum_{h=1}^J \exp(\mathbf{x}_i'\boldsymbol{\beta}_h)}
$$ {#eq-mlogit}

```{r}
#| label: mlogit-example
#| eval: false
library(nnet)
mlogit_model <- multinom(occupation ~ educ + exper + age, data = mydata)
```

::: {.callout-warning}
## IIA Assumption
Multinomial logit assumes **Independence of Irrelevant Alternatives** (IIA).

If violated, use:
- Nested logit
- Mixed logit
- Multinomial probit
:::

### Ordered Logit/Probit

For ordered outcomes (e.g., ratings: poor, fair, good, excellent):

```{r}
#| label: ologit-example
#| eval: false
library(MASS)
ologit_model <- polr(rating ~ income + age + education,
                     data = mydata, method = "logistic")
```

## Count Data Models {#sec-count-data}

### Poisson Regression

For count outcomes ($y \geq 0$):

$$
P(Y = y | \mathbf{x}) = \frac{e^{-\lambda} \lambda^y}{y!}, \quad \lambda = \exp(\mathbf{x}'\boldsymbol{\beta})
$$ {#eq-poisson}

```{r}
#| label: poisson-example
#| eval: false
poisson_model <- glm(num_arrests ~ age + educ + income,
                     family = poisson(link = "log"), data = mydata)
```

::: {.callout-warning}
## Overdispersion
Poisson assumes $\text{Var}(Y) = E(Y)$ (equidispersion).

If $\text{Var}(Y) > E(Y)$ (overdispersion), use:
- **Negative Binomial** regression
- Quasi-Poisson
:::

### Negative Binomial

Relaxes equidispersion assumption:

```{r}
#| label: negbin-example
#| eval: false
library(MASS)
negbin_model <- glm.nb(num_arrests ~ age + educ + income, data = mydata)
```

## Limited Dependent Variables {#sec-limited-dv}

### Tobit Model (Corner Solution)

For outcomes censored at zero (e.g., hours worked, expenditures):

$$
y_i^* = \mathbf{x}_i'\boldsymbol{\beta} + \epsilon_i, \quad y_i = \max(0, y_i^*)
$$ {#eq-tobit}

```{r}
#| label: tobit-example
#| eval: false
library(censReg)
tobit_model <- censReg(hours ~ educ + age + kids, data = mydata)
```

::: {.callout-note}
**Tobit** is appropriate when zeros represent corner solutions, not missing data.
:::

### Heckman Selection Model

For sample selection bias (non-random missing data):

**Selection equation**: $s_i^* = \mathbf{z}_i'\boldsymbol{\gamma} + u_i$, observe $s_i = 1(s_i^* > 0)$

**Outcome equation**: $y_i = \mathbf{x}_i'\boldsymbol{\beta} + \epsilon_i$, observe $y_i$ only if $s_i = 1$

```{r}
#| label: heckman-example
#| eval: false
library(sampleSelection)
heckman_model <- selection(selection = inlf ~ age + educ + kids,
                           outcome = log(wage) ~ educ + exper,
                           data = mydata)
```

::: {.callout-important}
## Heckman Two-Step vs ML
- **Two-step**: Faster, easier diagnostics
- **ML**: More efficient, joint estimation

Both require **exclusion restriction**: variable in selection but not outcome.
:::

## Model Comparison {#sec-model-comparison}

::: {.panel-tabset}

## Information Criteria

```{r}
#| label: ic-comparison
#| eval: false
AIC(logit_model, probit_model)
BIC(logit_model, probit_model)
```

## Likelihood Ratio Test

```{r}
#| label: lr-test
#| eval: false
lrtest(restricted_model, unrestricted_model)
```

## Vuong Test

For non-nested models:

```{r}
#| label: vuong-test
#| eval: false
library(pscl)
vuong(model1, model2)
```

:::

## Summary

Key takeaways for qualitative and limited DV models:

1. **LPM**: Simple but problematic (use for robustness checks)
2. **Logit/Probit**: Very similar in practice
3. **Report AME**: More interpretable than coefficients
4. **Count data**: Check for overdispersion
5. **Corner solutions**: Use Tobit
6. **Sample selection**: Use Heckman with exclusion restriction

::: {.callout-tip}
## Practical Advice
Always:
- Check model assumptions
- Report marginal effects
- Use robust standard errors
- Compare models with IC
:::

## Further Reading

- Greene (2018): *Econometric Analysis*
- Wooldridge (2010): *Econometric Analysis of Cross Section and Panel Data*
- [margins package](https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html)
