# Instrumental Variables and Two Stage Least Squares {#sec-iv}

```{r}
#| label: setup-iv
#| include: false
# Packages loaded in before-chapter.R
```

::: {.callout-note}
## Chapter Overview
This chapter covers instrumental variables (IV) estimation, a solution to endogeneity:

- The endogeneity problem and its sources
- Instrumental variable requirements
- Two-stage least squares (2SLS) estimation
- Testing for weak instruments and overidentification
- Panel IV estimation
- GMM estimation
:::

## The Endogeneity Problem {#sec-endogeneity}

### What is Endogeneity? {#sec-what-endogeneity}

In OLS regression, we assume that explanatory variables are **exogenous**:

$$E(\epsilon|X) = 0 \quad \text{or equivalently} \quad Cov(X_k, \epsilon) = 0$$

When this fails, we have **endogeneity**: $Cov(X_k, \epsilon) \neq 0$

::: {.callout-warning}
## Consequences of Endogeneity
- OLS is **biased**: $E(\hat{\beta}) \neq \beta$
- OLS is **inconsistent**: $\hat{\beta} \not\xrightarrow{p} \beta$
- Cannot interpret coefficients as causal effects
- **Solution**: Instrumental Variables (IV) or experiments
:::

### Sources of Endogeneity {#sec-sources-endogeneity}

::: {.panel-tabset}

## 1. Omitted Variables

**Problem**: Relevant variable left out of model

**Example**: Returns to education
- True model: $\log(wage) = \beta_0 + \beta_1 educ + \beta_2 ability + \epsilon$
- Estimated: $\log(wage) = \beta_0 + \beta_1 educ + u$
- Since $Cov(educ, ability) \neq 0$, education is endogenous

**Bias**: $E(\hat{\beta}_1) = \beta_1 + \beta_2 \frac{Cov(educ, ability)}{Var(educ)}$

## 2. Measurement Error

**Problem**: Variable measured with error

**Example**: True income vs reported income
- True model: $consumption = \beta_0 + \beta_1 income^* + \epsilon$
- Observed: $income = income^* + \nu$ (measurement error)
- This creates $Cov(income, u) \neq 0$ where $u = \epsilon - \beta_1\nu$

**Result**: **Attenuation bias** (coefficient biased toward zero)

## 3. Simultaneity

**Problem**: $X$ and $Y$ jointly determined (reverse causality)

**Example**: Supply and demand
- Price and quantity determined simultaneously
- $Q^d = \alpha_0 + \alpha_1 P + \epsilon_d$ (demand)
- $Q^s = \beta_0 + \beta_1 P + \epsilon_s$ (supply)
- Equilibrium: $Q^d = Q^s$
- Cannot identify elasticities from $(P, Q)$ data alone!

```{r}
#| label: fig-simultaneity
#| fig-cap: "Simultaneity: Price and Quantity jointly determined"
#| echo: false
#| fig-width: 7
#| fig-height: 5
library(ggplot2)

# Simulate supply and demand
P <- seq(0, 100, by = 1)
Qd1 <- 100 - 0.5 * P
Qs1 <- -20 + 0.8 * P
Qd2 <- 110 - 0.5 * P
Qs2 <- -15 + 0.8 * P

df <- data.frame(
  P = rep(P, 4),
  Q = c(Qd1, Qs1, Qd2, Qs2),
  Curve = rep(c("Demand 1", "Supply 1", "Demand 2", "Supply 2"), each = length(P))
)

ggplot(df, aes(x = Q, y = P, color = Curve, linetype = Curve)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Demand 1" = "blue", "Supply 1" = "red",
                                 "Demand 2" = "blue", "Supply 2" = "red")) +
  scale_linetype_manual(values = c("Demand 1" = "solid", "Supply 1" = "solid",
                                     "Demand 2" = "dashed", "Supply 2" = "dashed")) +
  labs(
    x = "Quantity",
    y = "Price",
    title = "Supply and Demand Shifts",
    subtitle = "Observed (P,Q) cannot identify elasticities"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

:::

## Instrumental Variables {#sec-iv-solution}

### The IV Idea {#sec-iv-idea}

Consider the simple model with endogenous $X$:

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i, \quad Cov(X_i, \epsilon_i) \neq 0
$$

**Solution**: Find an **instrumental variable** $Z$ that satisfies:

::: {.callout-important}
## Requirements for a Valid Instrument

**1. Independence (Exogeneity)**
$$Cov(Z, \epsilon) = 0$$
The instrument is **uncorrelated with the error term**

**2. Relevance**
$$Cov(Z, X) \neq 0$$
The instrument is **correlated with the endogenous variable**

These are called the **instrument validity conditions**.
:::

### Intuition {#sec-iv-intuition}

The IV estimator uses only the variation in $X$ that is **induced by $Z$**:

1. **OLS problem**: Captures correlation between $X$ and $Y$, including:
   - Direct effect: $X \rightarrow Y$ (what we want)
   - Indirect effect: $X \leftarrow \epsilon \rightarrow Y$ (bias!)

2. **IV solution**:
   - $Z$ affects $X$: $Z \rightarrow X$ (relevance)
   - $Z$ doesn't affect $\epsilon$: $Z \not\rightarrow \epsilon$ (exogeneity)
   - So variation in $X$ from $Z$ is **exogenous**!

3. **IV captures**: $Corr(Z \rightarrow X, Y)$ = causal effect of $X$ on $Y$

```{r}
#| label: fig-iv-dag
#| fig-cap: "Causal Diagram: IV breaks the endogeneity path"
#| echo: false
#| eval: false
# Could include DAG diagram here if dagitty/ggdag packages available
```

::: {.callout-tip}
## Classic IV Examples

**Returns to education**
- Endogenous: Years of education (correlated with ability)
- Instrument: Distance to college, quarter of birth
- Logic: Distance affects education but not ability/wage directly

**Effect of police on crime**
- Endogenous: Number of police (cities hire more police when crime is high)
- Instrument: Firefighters per capita, election timing
- Logic: Firefighters correlate with city resources but don't directly affect crime

**Effect of military service on earnings**
- Endogenous: Veteran status (selection into military)
- Instrument: Draft lottery number (Vietnam era)
- Logic: Random lottery affects service but not inherent earning potential
:::

## Two-Stage Least Squares (2SLS) {#sec-2sls}

### The 2SLS Procedure {#sec-2sls-procedure}

Consider the model:

$$
Y_i = \beta_0 + \beta_1 X_i + \beta_2 W_i + \epsilon_i
$$

Where:
- $X_i$ is endogenous: $Cov(X_i, \epsilon_i) \neq 0$
- $W_i$ is exogenous: $Cov(W_i, \epsilon_i) = 0$
- $Z_i$ is an instrumental variable for $X_i$

::: {.panel-tabset}

## Stage 1: First Stage

Regress endogenous variable on **all instruments and exogenous variables**:

$$
X_i = \pi_0 + \pi_1 Z_i + \pi_2 W_i + \nu_i
$$

Get predicted values: $\hat{X}_i = \hat{\pi}_0 + \hat{\pi}_1 Z_i + \hat{\pi}_2 W_i$

**Key test**: Is $\hat{\pi}_1$ significantly different from zero?
- If not, **weak instrument problem**!
- Rule of thumb: First-stage F-statistic > 10

## Stage 2: Second Stage

Regress $Y$ on predicted $\hat{X}$ and exogenous variables:

$$
Y_i = \beta_0 + \beta_1 \hat{X}_i + \beta_2 W_i + error_i
$$

This gives consistent estimates of $\beta_1$!

**Important**: Standard errors from naive 2SLS are **wrong**
- Must use proper IV standard errors
- Or use built-in IV functions (e.g., `iv_robust()`, `ivreg()`)

:::

### Simulation Example {#sec-2sls-simulation}

Let's simulate data to see IV in action:

```{r}
#| label: simulate-iv-data
library(fabricatr)
library(estimatr)

set.seed(123)
dat <- fabricate(
  N = 10000,
  # Unobserved confounder to create endogeneity
  U = rnorm(N),
  # Instrument (binary)
  Z1 = rbinom(N, 1, prob = 0.4),
  Z2 = rbinom(N, 1, prob = 0.4),
  # Exogenous variable
  X2 = rnorm(N),
  # Endogenous variable (correlated with U)
  X1 = 2 + 0.5 * Z1 + 0.5 * Z2 + 0.3 * X2 + U + rnorm(N, sd = 2),
  # Outcome (also affected by U, creating endogeneity in the Y ~ X1 relationship)
  Y = 3 + 0.8 * X1 + 0.5 * X2 + U + rnorm(N, sd = 1)
)

head(dat)
```

```{r}
#| label: compare-ols-iv
library(modelsummary)

# OLS (biased due to endogeneity)
ols_model <- lm(Y ~ X1 + X2, data = dat)

# Manual 2SLS (for illustration)
# Stage 1: Regress X1 on instruments and exogenous variables
first_stage <- lm(X1 ~ Z1 + Z2 + X2, data = dat)
X1_hat <- fitted(first_stage)

# Stage 2: Regress Y on predicted X1
second_stage <- lm(Y ~ X1_hat + X2, data = dat)

# Proper IV estimation (correct SEs)
iv_one <- iv_robust(Y ~ X1 + X2 | Z1 + X2, data = dat, diagnostics = TRUE)
iv_two <- iv_robust(Y ~ X1 + X2 | Z1 + Z2 + X2, data = dat, diagnostics = TRUE)

# Compare
modelsummary(
  list("OLS" = ols_model,
       "2SLS (manual)" = second_stage,
       "IV (1 instrument)" = iv_one,
       "IV (2 instruments)" = iv_two),
  stars = TRUE,
  gof_map = c("nobs", "r.squared")
)
```

::: {.callout-note}
## Interpreting Results
- **OLS**: Biased estimate of $\beta_1$ (endogeneity)
- **Manual 2SLS**: Correct point estimate, but SEs are wrong!
- **IV (proper)**: Consistent estimate with correct standard errors
- **More instruments**: Can improve efficiency (smaller SE)
:::

## Testing and Diagnostics {#sec-iv-diagnostics}

### Testing for Weak Instruments {#sec-weak-instruments}

**Problem**: If $Cov(Z, X)$ is weak, IV estimator is:
- Biased in small samples (bias toward OLS)
- Large standard errors
- Poor performance

::: {.callout-warning}
## Weak Instrument Problem
**Weak instruments** occur when first-stage F-statistic < 10

**Consequences**:
- IV estimates unreliable
- Hypothesis tests have wrong size
- Confidence intervals have wrong coverage

**Rule of thumb**: First-stage F > 10 (Stock and Yogo, 2005)
- F > 10: Reasonably strong
- F > 20: Strong instruments
- F < 10: Weak instruments (be very careful!)
:::

```{r}
#| label: weak-instrument-test
# Check first-stage F-statistic
summary(iv_one, diagnostics = TRUE)

# Extract F-stat
cat("First-stage F-statistic:", iv_one$diagnostic_first_stage_fstatistic[1], "\n")

# Manual calculation
first_stage <- lm(X1 ~ Z1 + X2, data = dat)
fs_summary <- summary(first_stage)

# F-test for instruments only (Z1 in this case)
restricted <- lm(X1 ~ X2, data = dat)
anova_test <- anova(restricted, first_stage)
cat("Manual F-statistic:", anova_test$F[2], "\n")
```

### Testing for Overidentification {#sec-overid-test}

When you have **more instruments than endogenous variables** (overidentified):

**Sargan-Hansen J test**:
- $H_0$: All instruments are valid (exogenous)
- $H_1$: At least one instrument is invalid

::: {.callout-note}
## Identification Status
- **Exactly identified**: # instruments = # endogenous variables
  - Cannot test instrument validity
- **Overidentified**: # instruments > # endogenous variables
  - Can test with Sargan/Hansen J test
  - More efficient estimation
- **Underidentified**: # instruments < # endogenous variables
  - Cannot estimate model!
:::

```{r}
#| label: overid-test
# With 2 instruments for 1 endogenous variable, we're overidentified
summary(iv_two, diagnostics = TRUE)

# J-statistic for overidentification
cat("J-statistic:", iv_two$diagnostic_overid_test, "\n")
cat("p-value:", iv_two$diagnostic_overid_test_p, "\n")
```

**Interpretation**:
- Large p-value → Cannot reject $H_0$ → Instruments appear valid
- Small p-value → Reject $H_0$ → At least one instrument likely invalid

### Testing for Endogeneity {#sec-endogeneity-test}

**Durbin-Wu-Hausman test**: Is $X$ actually endogenous?

- $H_0$: $X$ is exogenous (OLS is consistent and efficient)
- $H_1$: $X$ is endogenous (need IV)

```{r}
#| label: endogeneity-test
#| eval: false
library(AER)

# Hausman test
ivreg_model <- ivreg(Y ~ X1 + X2 | Z1 + Z2 + X2, data = dat)
summary(ivreg_model, diagnostics = TRUE)
```

**If $H_0$ not rejected**: Use OLS (more efficient)
**If $H_0$ rejected**: Use IV (consistent despite inefficiency)

## Real Data Example {#sec-iv-example}

### Card (1995): Returns to Education {#sec-card-example}

**Research question**: What is the causal effect of education on wages?

**Endogeneity**: $Cov(educ, \epsilon) \neq 0$ due to omitted ability

**Instrument**: Proximity to 4-year college
- Affects education (relevance)
- Doesn't directly affect ability/wages (exogeneity)

```{r}
#| label: card-data
# Load Card (1995) data
library(AER)
data("CollegeDistance", package = "AER")

# Variables:
# wage: log wage
# education: years of education
# distance: distance to nearest 4-year college (miles/10)
# Controls: gender, ethnicity, urban, unemp, tuition, etc.

# Summary
datasummary_skim(CollegeDistance[c("wage", "education", "distance")])
```

::: {.panel-tabset}

## OLS (Biased)

```{r}
#| label: card-ols
ols_card <- lm(wage ~ education + gender + ethnicity + unemp + urban,
               data = CollegeDistance)
summary(ols_card)
```

**Return to education** (OLS): `r round(100 * coef(ols_card)["education"], 2)`%

## IV (Consistent)

```{r}
#| label: card-iv
iv_card <- ivreg(wage ~ education + gender + ethnicity + unemp + urban |
                   distance + gender + ethnicity + unemp + urban,
                 data = CollegeDistance)
summary(iv_card, diagnostics = TRUE)
```

**Return to education** (IV): `r round(100 * coef(iv_card)["education"], 2)`%

## Comparison

```{r}
#| label: card-compare
modelsummary(
  list("OLS" = ols_card, "IV" = iv_card),
  stars = TRUE,
  gof_map = c("nobs", "r.squared")
)
```

:::

**Key findings**:
- IV estimate > OLS estimate
- Suggests **negative ability bias** (ability negatively correlated with distance)
- First-stage F = `r round(summary(iv_card, diagnostics = TRUE)$diagnostics["Weak instruments", "statistic"], 1)` (> 10, strong instrument)

## Robust Standard Errors {#sec-iv-robust}

Just like OLS, IV estimates can have heteroskedasticity:

```{r}
#| label: iv-robust-se
library(sandwich)
library(lmtest)

# IV with robust standard errors
coeftest(iv_card, vcov = vcovHC(iv_card, type = "HC1"))
```

Or use `iv_robust()` which gives robust SEs by default:

```{r}
#| label: iv-robust-estimatr
iv_robust_card <- iv_robust(
  wage ~ education + gender + ethnicity + unemp + urban |
    distance + gender + ethnicity + unemp + urban,
  data = CollegeDistance
)
summary(iv_robust_card)
```

## Panel IV {#sec-panel-iv}

### Fixed Effects IV {#sec-fe-iv}

With panel data, combine IV with fixed effects:

$$
Y_{it} = \beta_1 X_{it} + \alpha_i + \lambda_t + \epsilon_{it}
$$

Where $X_{it}$ is endogenous, $\alpha_i$ are individual fixed effects.

**Instrument**: $Z_{it}$ that varies over time within individuals

```{r}
#| label: panel-iv-example
#| eval: false
library(plm)

# Panel IV with fixed effects
panel_iv <- plm(
  y ~ x1 + x2 | z1 + z2,
  data = panel_data,
  model = "within",  # Fixed effects
  index = c("id", "time")
)
summary(panel_iv)
```

### Using fixest for Panel IV {#sec-fixest-iv}

The `fixest` package provides fast and flexible panel IV estimation:

```{r}
#| label: fixest-iv
#| eval: false
library(fixest)

# Panel IV with individual and time fixed effects
feiv_model <- feols(
  y ~ x2 | id + time | x1 ~ z1 + z2,
  data = panel_data,
  vcov = "hetero"  # Robust SEs
)
summary(feiv_model)
```

**Syntax**: `feols(outcome ~ exog_vars | fixed_effects | endog ~ instruments)`

## GMM Estimation {#sec-gmm}

### Generalized Method of Moments {#sec-gmm-intro}

**GMM** is a general estimation framework that includes IV as a special case.

**Moment conditions**: $E[Z_i(Y_i - X_i'\beta)] = 0$

With **more instruments than parameters** (overidentified):
- Can estimate efficiently
- Can test overidentifying restrictions

```{r}
#| label: gmm-example
#| eval: false
library(gmm)

# GMM IV estimation
g <- function(theta, data) {
  Y <- data[, 1]
  X <- data[, 2:3]
  Z <- data[, 4:6]

  # Moment conditions: E[Z'(Y - X'β)] = 0
  moments <- Z * (Y - X %*% theta)
  return(moments)
}

# Estimate
gmm_result <- gmm(
  g,
  x = cbind(dat$Y, dat$X1, dat$X2, dat$Z1, dat$Z2, dat$X2),
  t0 = c(1, 1)  # Starting values
)
summary(gmm_result)
```

::: {.callout-note}
## When to Use GMM?
- **Overidentified models**: Efficient estimation
- **Complex moment conditions**: Nonlinear models
- **Heteroskedasticity/autocorrelation**: GMM automatically robust

**Standard IV = GMM** with orthogonality conditions
:::

## Local Average Treatment Effect (LATE) {#sec-late}

### What IV Actually Estimates {#sec-late-interpretation}

IV doesn't estimate the **average treatment effect** (ATE)!

Instead, IV estimates the **local average treatment effect** (LATE):
- Effect for **compliers** only
- Compliers: Those whose treatment status is affected by the instrument

::: {.callout-important}
## LATE Framework (Imbens-Angrist, 1994)

**Types of individuals**:
1. **Compliers**: $D_i(Z=1) = 1, D_i(Z=0) = 0$ (respond to instrument)
2. **Always-takers**: $D_i(Z=1) = 1, D_i(Z=0) = 1$ (always treated)
3. **Never-takers**: $D_i(Z=1) = 0, D_i(Z=0) = 0$ (never treated)
4. **Defiers**: $D_i(Z=1) = 0, D_i(Z=0) = 1$ (do opposite)

**IV identifies**: Effect for compliers only!

$$\beta_{IV} = E[Y_i(1) - Y_i(0) | \text{Complier}]$$

**Not**: $E[Y_i(1) - Y_i(0)]$ (average treatment effect)
:::

**Example**: Draft lottery and veteran earnings
- Compliers: Drafted → served, Not drafted → didn't serve
- Always-takers: Volunteered regardless
- Never-takers: Avoided service regardless
- IV = Effect for those whose service was affected by draft

## Common Mistakes {#sec-iv-mistakes}

::: {.callout-warning}
## Pitfalls to Avoid

**1. Using naive 2SLS standard errors**
- Manual 2SLS gives wrong SEs!
- Always use proper IV functions

**2. Weak instruments**
- First-stage F < 10 → unreliable estimates
- Report first-stage results!

**3. Invalid instruments**
- Exogeneity cannot be tested directly
- Requires strong economic reasoning
- Use overidentification tests when possible

**4. Misinterpreting IV**
- IV ≠ ATE (it's LATE)
- Local to compliers only
- Different instruments → different LATEs

**5. Forgetting reduced form**
- Always report: First stage + Reduced form + IV
- Helps assess instrument validity
:::

## Summary {#sec-iv-summary}

::: {.callout-note}
## Key Takeaways

**Endogeneity**:
- $Cov(X, \epsilon) \neq 0$ → OLS biased and inconsistent
- Sources: Omitted variables, measurement error, simultaneity

**Instrumental Variables**:
- Must be exogenous: $Cov(Z, \epsilon) = 0$
- Must be relevant: $Cov(Z, X) \neq 0$
- Finding good instruments is hard!

**2SLS Estimation**:
- Stage 1: Regress $X$ on $Z$ (and other exogenous variables)
- Stage 2: Regress $Y$ on $\hat{X}$
- Use proper IV functions for correct SEs

**Diagnostics**:
- First-stage F > 10 (weak instruments)
- Sargan-Hansen test (overidentification)
- Hausman test (endogeneity)

**Interpretation**:
- IV estimates LATE (local average treatment effect)
- Effect for compliers, not population average
- Larger SEs than OLS (price of consistency)

**Best Practices**:
- Report first-stage results
- Test for weak instruments
- Provide economic justification for instruments
- Consider LATE interpretation
:::

## Further Reading {#sec-iv-references}

- @wooldridge2015introductory - Chapter 15
- @angrist2009mostly - Chapters 4-6 (intuitive)
- @imbens1994identification - LATE framework
- @stock2005testing - Weak instruments
- @murray2006testing - Weak instrument robust inference

---

::: {.callout-tip}
## Next Chapter
See @sec-panel for **Panel Data Models** and @sec-glm for **Limited Dependent Variables**.
:::
