# Linear Regression {#sec-lm}

```{r}
#| label: setup-lm
#| include: false
# Packages loaded in before-chapter.R
```

::: {.callout-note}
## Chapter Overview
This chapter covers linear regression, the workhorse of econometrics:

- Simple and multiple linear regression
- OLS estimation and properties
- Statistical inference (t-tests, F-tests, confidence intervals)
- Qualitative variables (dummy variables, interactions)
- Functional forms (logs, polynomials, splines)
- Heteroskedasticity testing and correction
- Model specification and selection
- Quantile regression
:::

## Simple Linear Regression {#sec-slr}

### The Model {#sec-slr-model}

We start with the simple OLS regression model relating a dependent variable $Y$ to a single explanatory variable $X$:

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
$$ {#eq-slr}

Where:

- $Y_i$: Dependent variable (e.g., salary)
- $X_i$: Independent variable (e.g., ROE - return on equity)
- $\beta_0$: Intercept
- $\beta_1$: Slope coefficient (marginal effect of $X$ on $Y$)
- $\epsilon_i$: Error term with $E(\epsilon_i|X_i) = 0$

::: {.callout-tip}
## Classic SLR Examples
- College GPA ~ SAT scores ($\beta > 0$)
- Change in GDP ~ change in unemployment ($\beta < 0$)
- House price ~ number of bedrooms ($\beta > 0$)
- Fatalities per year ~ speed limit ($\beta < 0$)

These are simplifications of more complex relationships.
:::

### Example: CEO Salary and ROE {#sec-ceo-example}

Let's analyze CEO compensation using the `ceosal1` dataset:

```{r}
#| label: load-ceosal1
data("ceosal1", package = "wooldridge")
```

```{r}
#| label: explore-ceosal1
#| echo: false
knitr::kable(
  head(ceosal1[,1:8], 10),
  caption = "CEO Salary Data (first 8 columns)"
)
```

```{r}
#| label: ceosal1-summary
# Data structure
str(ceosal1)

# Summary statistics
summary(ceosal1[c("salary", "roe")])
```

**Research Question**: How does return on equity (ROE) affect CEO salary?

**Expected sign**: We expect $\beta_1 > 0$ (higher firm performance → higher CEO pay)

```{r}
#| label: fig-ceosal1
#| fig-cap: "Relationship between ROE and Salary"
#| fig-width: 7
#| fig-height: 5
library(ggplot2)
ggplot(ceosal1, aes(x = roe, y = salary)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "Return on Equity (%)",
    y = "Salary ($1000s)",
    title = "CEO Salary vs. Firm Performance"
  ) +
  theme_minimal()
```

### OLS Estimation {#sec-ols-slr}

The OLS estimators minimize the sum of squared residuals:

$$
\min_{\beta_0, \beta_1} \sum_{i=1}^n (Y_i - \beta_0 - \beta_1 X_i)^2
$$

**OLS formulas**:

$$
\hat{\beta}_1 = \frac{\sum(X_i - \bar{X})(Y_i - \bar{Y})}{\sum(X_i - \bar{X})^2} = \frac{\text{Cov}(X,Y)}{\text{Var}(X)}
$$ {#eq-beta1-slr}

$$
\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1 \bar{X}
$$ {#eq-beta0-slr}

::: {.panel-tabset}

## Manual Calculation

```{r}
#| label: manual-ols
# Ingredients
cov_xy <- cov(ceosal1$roe, ceosal1$salary)
var_x <- var(ceosal1$roe)
mean_y <- mean(ceosal1$salary)
mean_x <- mean(ceosal1$roe)

# OLS estimates
b1hat <- cov_xy / var_x
b0hat <- mean_y - b1hat * mean_x

cat("Manual OLS estimates:\n")
cat(sprintf("β₀ = %.2f\n", b0hat))
cat(sprintf("β₁ = %.2f\n", b1hat))
```

## Using lm()

```{r}
#| label: lm-ceosal1
# OLS regression
model1 <- lm(salary ~ roe, data = ceosal1)
summary(model1)
```

:::

**Interpretation**: A one percentage point increase in ROE is associated with an increase in CEO salary of approximately $`r round(coef(model1)[2], 2)` thousand dollars.

```{r}
#| label: fig-ols-fit
#| fig-cap: "OLS Regression Line"
#| fig-width: 7
#| fig-height: 5
ggplot(ceosal1, aes(x = roe, y = salary)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    x = "Return on Equity (%)",
    y = "Salary ($1000s)",
    title = "CEO Salary vs. ROE with OLS Fit"
  ) +
  theme_minimal()
```

### OLS Assumptions {#sec-slr-assumptions}

::: {.callout-important}
## Key Assumptions for SLR

**SLR.1**: **Linear in parameters**
$$Y = \beta_0 + \beta_1 X + \epsilon$$

**SLR.2**: **Random sampling**
$(X_i, Y_i)$ are independent and identically distributed

**SLR.3**: **Sample variation in $X$**
$\text{Var}(X) > 0$ (not all $X_i$ are the same)

**SLR.4**: **Zero conditional mean**
$$E(\epsilon|X) = 0$$
This is the key assumption! It means $X$ is exogenous.

**SLR.5**: **Homoskedasticity** (for inference)
$$\text{Var}(\epsilon|X) = \sigma^2$$
Constant error variance

**SLR.6**: **Normality** (for exact inference in small samples)
$$\epsilon \sim N(0, \sigma^2)$$
:::

**Under SLR.1-SLR.4**:
- OLS is **unbiased**: $E(\hat{\beta}_1) = \beta_1$
- OLS is **consistent**: $\hat{\beta}_1 \xrightarrow{p} \beta_1$

**Under SLR.1-SLR.5** (Gauss-Markov):
- OLS is **BLUE**: Best Linear Unbiased Estimator
- Has smallest variance among all linear unbiased estimators

### Properties and Residuals {#sec-slr-properties}

```{r}
#| label: ols-properties
# Extract key quantities
names(model1)

# Coefficients
coef(model1)

# Residuals
head(resid(model1))

# Fitted values
head(fitted(model1))

# Sum of residuals (always zero!)
sum(resid(model1))

# Correlation between X and residuals (always zero!)
cor(ceosal1$roe, resid(model1))
```

::: {.callout-note}
## OLS Properties
- Residuals sum to zero: $\sum \hat{\epsilon}_i = 0$
- Residuals uncorrelated with $X$: $\sum X_i\hat{\epsilon}_i = 0$
- Regression line passes through $(\bar{X}, \bar{Y})$
:::

### Goodness of Fit {#sec-r-squared}

$$
R^2 = \frac{SSE}{SST} = 1 - \frac{SSR}{SST}
$$ {#eq-r2}

Where:
- $SST = \sum(Y_i - \bar{Y})^2$ (Total Sum of Squares)
- $SSE = \sum(\hat{Y}_i - \bar{Y})^2$ (Explained Sum of Squares)
- $SSR = \sum\hat{\epsilon}_i^2$ (Residual Sum of Squares)

```{r}
#| label: r-squared
# R-squared from summary
summary(model1)$r.squared

# Manual calculation
y_bar <- mean(ceosal1$salary)
SST <- sum((ceosal1$salary - y_bar)^2)
SSR <- sum(resid(model1)^2)
R2 <- 1 - SSR/SST
cat(sprintf("R² = %.4f\n", R2))
```

::: {.callout-warning}
## Interpreting R²
$R^2 = `r round(summary(model1)$r.squared, 3)`$ means ROE explains only `r round(summary(model1)$r.squared * 100, 1)`% of the variation in CEO salaries.

**Low R² doesn't mean the model is bad!** What matters is:
- Are coefficients statistically significant?
- Do they have expected signs?
- Is the model useful for our research question?
:::

## Multiple Linear Regression {#sec-mlr}

### The MLR Model {#sec-mlr-model}

With $k$ explanatory variables:

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + \epsilon_i
$$ {#eq-mlr}

**Matrix form**:

$$
\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$ {#eq-mlr-matrix}

**OLS estimator**:

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}
$$ {#eq-ols-matrix}

### MLR Assumptions {#sec-mlr-assumptions}

::: {.callout-important}
## Gauss-Markov Assumptions (MLR)

**MLR.1**: **Linear in parameters**

**MLR.2**: **Random sampling**

**MLR.3**: **No perfect multicollinearity**
No $X$ variable is constant, and no exact linear relationship among $X$'s

**MLR.4**: **Zero conditional mean**
$$E(\epsilon|\mathbf{X}) = 0$$
All $X$'s are exogenous

**MLR.5**: **Homoskedasticity**
$$\text{Var}(\epsilon|\mathbf{X}) = \sigma^2$$

**Under MLR.1-MLR.5**: OLS is **BLUE** (Best Linear Unbiased Estimator)
:::

### Example: College GPA {#sec-gpa-example}

```{r}
#| label: gpa-data
data("gpa1", package = "wooldridge")

# Multiple regression
gpa_model <- lm(colGPA ~ hsGPA + ACT + skipped, data = gpa1)
summary(gpa_model)
```

**Interpretation**:
- Holding HS GPA and ACT constant, each additional skipped lecture reduces college GPA by `r abs(round(coef(gpa_model)["skipped"], 3))` points
- **Ceteris paribus** (all else equal) interpretation is key!

::: {.panel-tabset}

## Simple vs Multiple

```{r}
#| label: compare-models
# Simple regression (omitted variable bias?)
simple <- lm(colGPA ~ skipped, data = gpa1)

# Multiple regression
multiple <- lm(colGPA ~ hsGPA + ACT + skipped, data = gpa1)

# Compare coefficients on 'skipped'
library(modelsummary)
modelsummary(
  list("Simple" = simple, "Multiple" = multiple),
  stars = TRUE,
  gof_map = c("nobs", "r.squared")
)
```

## Visualizing MLR

```{r}
#| label: fig-mlr-viz
#| fig-cap: "Partial Regression Plot: College GPA vs Skipped Classes"
#| fig-width: 7
#| fig-height: 5
library(car)
avPlots(gpa_model, terms = ~ skipped)
```

:::

### Partialling Out Interpretation {#sec-partialling-out}

The coefficient $\hat{\beta}_1$ in:
$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k + \epsilon$$

Can be obtained by:

1. Regress $Y$ on $X_2, ..., X_k$ and get residuals $\tilde{Y}$
2. Regress $X_1$ on $X_2, ..., X_k$ and get residuals $\tilde{X}_1$
3. Regress $\tilde{Y}$ on $\tilde{X}_1$ (no intercept)

This is called **Frisch-Waugh-Lovell theorem** - $\hat{\beta}_1$ captures the effect of $X_1$ on $Y$ **after removing the effects of other $X$'s**.

## Statistical Inference {#sec-inference}

### Standard Errors {#sec-standard-errors}

Under MLR.1-MLR.5, the variance of $\hat{\beta}_j$ is:

$$
\text{Var}(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)}
$$ {#eq-var-beta}

Where:
- $SST_j = \sum(X_{ji} - \bar{X}_j)^2$: variation in $X_j$
- $R_j^2$: R² from regressing $X_j$ on all other $X$'s

**Standard error**:

$$
\text{SE}(\hat{\beta}_j) = \sqrt{\frac{\hat{\sigma}^2}{SST_j(1-R_j^2)}}
$$ {#eq-se}

Where $\hat{\sigma}^2 = \frac{SSR}{n-k-1}$ (residual variance)

::: {.callout-note}
## What Affects SE?
- **Larger sample** ($n \uparrow$) → SE $\downarrow$ (good!)
- **More variation in $X_j$** ($SST_j \uparrow$) → SE $\downarrow$ (good!)
- **High multicollinearity** ($R_j^2 \uparrow$) → SE $\uparrow$ (bad!)
- **Larger errors** ($\sigma^2 \uparrow$) → SE $\uparrow$ (bad!)
:::

### t-Tests {#sec-t-tests}

Test $H_0: \beta_j = 0$ vs $H_1: \beta_j \neq 0$:

$$
t = \frac{\hat{\beta}_j - 0}{\text{SE}(\hat{\beta}_j)} \sim t_{n-k-1}
$$ {#eq-t-stat}

**Decision rule**:
- Reject $H_0$ if $|t| > t_{n-k-1,\alpha/2}$ (two-sided test)
- Or equivalently, reject if $p$-value $< \alpha$

**Confidence interval**:

$$
CI_{1-\alpha} = \hat{\beta}_j \pm t_{n-k-1,\alpha/2} \cdot \text{SE}(\hat{\beta}_j)
$$ {#eq-ci}

```{r}
#| label: t-tests
# t-statistics and p-values
summary(gpa_model)$coefficients

# Confidence intervals (95%)
confint(gpa_model, level = 0.95)
```

::: {.panel-tabset}

## One-Sided Test

```{r}
#| label: one-sided
# Test H0: β_skipped = 0 vs H1: β_skipped < 0
t_stat <- coef(summary(gpa_model))["skipped", "t value"]
p_value_one_sided <- pt(t_stat, df = gpa_model$df.residual)

cat(sprintf("t-statistic: %.3f\n", t_stat))
cat(sprintf("One-sided p-value: %.4f\n", p_value_one_sided))
```

## Two-Sided Test

The `summary()` output provides two-sided p-values by default.

```{r}
#| label: two-sided
coef(summary(gpa_model))["skipped", ]
```

:::

### F-Tests {#sec-f-tests}

Test **multiple restrictions** simultaneously:

$$H_0: \beta_1 = \beta_2 = \cdots = \beta_q = 0$$

$$
F = \frac{(SSR_r - SSR_{ur})/q}{SSR_{ur}/(n-k-1)} \sim F_{q,n-k-1}
$$ {#eq-f-stat}

Where:
- $SSR_r$: SSR from restricted model (under $H_0$)
- $SSR_{ur}$: SSR from unrestricted model
- $q$: number of restrictions

::: {.panel-tabset}

## Overall F-Test

Test $H_0: \beta_1 = \beta_2 = \cdots = \beta_k = 0$ (all slopes are zero)

```{r}
#| label: overall-f
summary(gpa_model)$fstatistic

# Manual calculation
R2 <- summary(gpa_model)$r.squared
n <- nobs(gpa_model)
k <- length(coef(gpa_model)) - 1

F_stat <- (R2 / k) / ((1 - R2) / (n - k - 1))
p_value <- 1 - pf(F_stat, k, n - k - 1)

cat(sprintf("F-statistic: %.2f\n", F_stat))
cat(sprintf("p-value: %.4e\n", p_value))
```

## Partial F-Test

Test if a **subset** of variables are jointly significant:

```{r}
#| label: partial-f
# Restricted model: colGPA ~ hsGPA only
restricted <- lm(colGPA ~ hsGPA, data = gpa1)

# Unrestricted model: colGPA ~ hsGPA + ACT + skipped
unrestricted <- gpa_model

# F-test
anova(restricted, unrestricted)
```

**Interpretation**: ACT and skipped are jointly significant ($p < 0.001$).

:::

### Linear Restrictions {#sec-linear-restrictions}

Test more general restrictions using **Wald test**:

$$H_0: \beta_j = a_j \text{ for some value } a_j$$

```{r}
#| label: linear-restrictions
library(car)

# Test H0: β_hsGPA = 0.5
linearHypothesis(gpa_model, "hsGPA = 0.5")

# Test H0: β_ACT = 0 AND β_skipped = 0
linearHypothesis(gpa_model, c("ACT = 0", "skipped = 0"))
```

## Qualitative Variables {#sec-dummy-variables}

### Dummy Variables {#sec-dummies}

A **dummy variable** (binary variable) takes values 0 or 1:

$$D_i = \begin{cases} 1 & \text{if condition is true} \\ 0 & \text{otherwise} \end{cases}$$

**Example**: Gender, race, treatment status, time period

::: {.callout-tip}
## Dummy Variable Trap
With $m$ categories, include only $m-1$ dummies!

Example: Male/Female
- Include `male` (0/1) OR `female` (0/1), not both
- Omitted category is the **reference group**
:::

### Example: Wage Equation {#sec-wage-example}

```{r}
#| label: wage-dummy
data("wage1", package = "wooldridge")

# Log wage regression with gender dummy
wage_model <- lm(log(wage) ~ educ + exper + I(exper^2) + female, data = wage1)
summary(wage_model)
```

**Interpretation**:
- `female` coefficient = `r round(coef(wage_model)["female"], 3)`
- Holding education and experience constant, women earn approximately `r abs(round(100 * coef(wage_model)["female"], 1))`% less than men
- This is the **gender wage gap** (ceteris paribus)

::: {.panel-tabset}

## Separate Intercepts

```{r}
#| label: fig-wage-gender
#| fig-cap: "Wage-Education Relationship by Gender"
#| fig-width: 7
#| fig-height: 5
ggplot(wage1, aes(x = educ, y = log(wage), color = factor(female))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Male", "Female"),
    name = "Gender"
  ) +
  labs(
    x = "Years of Education",
    y = "Log Wage",
    title = "Parallel regression lines (different intercepts)"
  ) +
  theme_minimal()
```

## Multiple Categories

```{r}
#| label: multiple-dummies
# Marital status (3 categories: never married, married, divorced/separated)
# Create dummies (married as reference)
wage1$never_married <- as.numeric(wage1$marr == 0 & wage1$single == 1)
wage1$divorced <- as.numeric(wage1$marr == 0 & wage1$single == 0)

marital_model <- lm(log(wage) ~ educ + exper + I(exper^2) +
                      never_married + divorced, data = wage1)
summary(marital_model)$coefficients[c("never_married", "divorced"), ]
```

**Interpretation** (relative to married):
- Never married: `r round(100 * coef(marital_model)["never_married"], 1)`% wage difference
- Divorced: `r round(100 * coef(marital_model)["divorced"], 1)`% wage difference

:::

### Interactions with Dummies {#sec-interactions}

**Interaction term**: $X \cdot D$ allows the effect of $X$ to differ across groups

$$
Y = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 (X \cdot D) + \epsilon
$$

- $\beta_1$: effect of $X$ when $D=0$ (reference group)
- $\beta_1 + \beta_3$: effect of $X$ when $D=1$
- $\beta_3$: **differential effect** of $X$ across groups

```{r}
#| label: interaction-model
# Allow education return to differ by gender
interact_model <- lm(log(wage) ~ educ + exper + I(exper^2) +
                       female + educ:female, data = wage1)
summary(interact_model)$coefficients[c("educ", "female", "educ:female"), ]
```

**Interpretation**:
- Return to education for men: `r round(100 * coef(interact_model)["educ"], 2)`%
- Return to education for women: `r round(100 * (coef(interact_model)["educ"] + coef(interact_model)["educ:female"]), 2)`%
- Difference: `r round(100 * coef(interact_model)["educ:female"], 2)`% (not statistically significant at 5% level)

```{r}
#| label: fig-interaction
#| fig-cap: "Wage-Education with Gender Interaction"
#| fig-width: 7
#| fig-height: 5
ggplot(wage1, aes(x = educ, y = log(wage), color = factor(female))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Male", "Female"),
    name = "Gender"
  ) +
  labs(
    x = "Years of Education",
    y = "Log Wage",
    title = "Different slopes (interaction effect)"
  ) +
  theme_minimal()
```

### Seasonal Dummies {#sec-seasonal}

With quarterly or monthly data, include seasonal dummies:

$$
Y_t = \beta_0 + \beta_1 X_t + \beta_2 Q2_t + \beta_3 Q3_t + \beta_4 Q4_t + \epsilon_t
$$

Omit Q1 as reference. Similar for monthly dummies (omit one month).

## Functional Forms {#sec-functional-forms}

### Log Transformations {#sec-logs}

::: {.panel-tabset}

## Level-Level

$$Y = \beta_0 + \beta_1 X + \epsilon$$

**Interpretation**: $\Delta Y = \beta_1 \Delta X$

One unit increase in $X$ → $\beta_1$ unit increase in $Y$

## Level-Log

$$Y = \beta_0 + \beta_1 \log(X) + \epsilon$$

**Interpretation**: $\Delta Y = (\beta_1/100) \%\Delta X$

1% increase in $X$ → $\beta_1/100$ unit increase in $Y$

## Log-Level

$$\log(Y) = \beta_0 + \beta_1 X + \epsilon$$

**Interpretation**: $\%\Delta Y = 100\beta_1 \Delta X$

One unit increase in $X$ → $100\beta_1$% increase in $Y$

## Log-Log (Elasticity)

$$\log(Y) = \beta_0 + \beta_1 \log(X) + \epsilon$$

**Interpretation**: $\%\Delta Y = \beta_1 \%\Delta X$

$\beta_1$ is the **elasticity**: 1% increase in $X$ → $\beta_1$% increase in $Y$

:::

**Example**: Housing prices

```{r}
#| label: hprice-example
data("hprice1", package = "wooldridge")

# Different functional forms
level_level <- lm(price ~ sqrft + bdrms, data = hprice1)
level_log <- lm(price ~ log(sqrft) + bdrms, data = hprice1)
log_level <- lm(log(price) ~ sqrft + bdrms, data = hprice1)
log_log <- lm(log(price) ~ log(sqrft) + bdrms, data = hprice1)

modelsummary(
  list("Level-Level" = level_level,
       "Level-Log" = level_log,
       "Log-Level" = log_level,
       "Log-Log" = log_log),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)
```

::: {.callout-note}
## Choosing Functional Form
- **Theory**: What does economic theory suggest?
- **Data**: Plot the relationship
- **Model fit**: Compare R², information criteria
- **Interpretation**: Which makes most sense?
:::

### Polynomials {#sec-polynomials}

Capture **nonlinear** relationships:

$$
Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon
$$

**Marginal effect**:

$$
\frac{\partial Y}{\partial X} = \beta_1 + 2\beta_2 X
$$

Depends on level of $X$!

```{r}
#| label: polynomial-example
# Wage-experience relationship (nonlinear)
wage_poly <- lm(log(wage) ~ educ + exper + I(exper^2), data = wage1)
summary(wage_poly)
```

**Interpretation**:
- Return to experience varies with experience level
- Peak at $X^* = -\beta_1/(2\beta_2) = `r round(-coef(wage_poly)["exper"]/(2*coef(wage_poly)["I(exper^2)"]), 1)`$ years

```{r}
#| label: fig-polynomial
#| fig-cap: "Nonlinear Wage-Experience Profile"
#| fig-width: 7
#| fig-height: 5
# Create prediction data
pred_data <- data.frame(
  exper = seq(0, 50, by = 1),
  educ = mean(wage1$educ)
)
pred_data$pred_wage <- exp(predict(wage_poly, newdata = pred_data))

ggplot(pred_data, aes(x = exper, y = pred_wage)) +
  geom_line(color = "darkblue", linewidth = 1) +
  labs(
    x = "Years of Experience",
    y = "Predicted Wage",
    title = "Quadratic wage-experience profile"
  ) +
  theme_minimal()
```

### Piecewise Linear (Splines) {#sec-splines}

Allow different slopes in different regions:

$$
Y = \beta_0 + \beta_1 X + \beta_2 (X - c) \cdot \mathbb{1}(X > c) + \epsilon
$$

Where $c$ is the **knot** (threshold).

```{r}
#| label: spline-example
#| eval: false
# Experience with knot at 10 years
wage1$exper_10plus <- pmax(wage1$exper - 10, 0)

spline_model <- lm(log(wage) ~ educ + exper + exper_10plus, data = wage1)
summary(spline_model)
```

## Heteroskedasticity {#sec-heteroskedasticity}

### The Problem {#sec-hetero-problem}

**Heteroskedasticity**: $\text{Var}(\epsilon_i|X_i) = \sigma_i^2$ varies with $X$

::: {.callout-warning}
## Consequences of Heteroskedasticity
- OLS estimators remain **unbiased** and **consistent**
- Standard errors are **biased** (usually too small)
- t-tests and F-tests are **invalid**
- OLS is **no longer BLUE** (inefficient)

**Solution**: Use **robust standard errors** or **weighted least squares**
:::

### Testing for Heteroskedasticity {#sec-hetero-tests}

::: {.panel-tabset}

## Breusch-Pagan Test

Test $H_0: \text{Var}(\epsilon|X) = \sigma^2$ (homoskedasticity)

```{r}
#| label: bp-test
library(lmtest)

# Breusch-Pagan test
bptest(wage_model)
```

## White Test

More general test (allows for arbitrary heteroskedasticity):

```{r}
#| label: white-test
# White test (includes squares and cross-products)
bptest(wage_model, ~ educ*exper + I(educ^2) + I(exper^2), data = wage1)
```

## Visual Inspection

```{r}
#| label: fig-hetero
#| fig-cap: "Residuals vs Fitted (checking heteroskedasticity)"
#| fig-width: 7
#| fig-height: 5
plot(wage_model, which = 1)
```

Look for **funnel shape** or other patterns.

:::

### Robust Standard Errors {#sec-robust-se}

**Heteroskedasticity-robust** (HC) standard errors:

```{r}
#| label: robust-se
library(sandwich)

# OLS with robust SEs (HC1)
coeftest(wage_model, vcov = vcovHC(wage_model, type = "HC1"))

# Compare to default SEs
coeftest(wage_model)
```

::: {.callout-tip}
## Types of Robust SEs
- **HC0**: Basic White robust SEs
- **HC1**: Degrees of freedom correction (recommended)
- **HC2**, **HC3**: More sophisticated (for leverage points)

Most papers use **HC1** or **HC3**.
:::

### Weighted Least Squares {#sec-wls}

If we know the form of heteroskedasticity: $\text{Var}(\epsilon_i|X_i) = \sigma^2 h_i$

**WLS**: Weight observations by $w_i = 1/\sqrt{h_i}$

$$
\min \sum_{i=1}^n w_i (Y_i - \beta_0 - \beta_1 X_{1i} - \cdots)^2
$$

```{r}
#| label: wls-example
#| eval: false
# If Var(ε) proportional to X²
weights <- 1 / wage1$educ^2

wls_model <- lm(log(wage) ~ educ + exper + I(exper^2),
                data = wage1, weights = weights)
summary(wls_model)
```

WLS is **more efficient** than OLS when heteroskedasticity form is correctly specified.

## Model Specification {#sec-specification}

### Omitted Variable Bias {#sec-ovb}

Omitting a relevant variable causes **bias**:

True model: $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \epsilon$

Estimated model: $Y = \beta_0 + \beta_1 X_1 + error$

**Bias**:

$$
E(\tilde{\beta}_1) = \beta_1 + \beta_2 \frac{\text{Cov}(X_1, X_2)}{\text{Var}(X_1)}
$$ {#eq-ovb}

::: {.callout-important}
## When is OVB a Problem?
OVB occurs when BOTH:
1. Omitted variable $X_2$ affects $Y$ ($\beta_2 \neq 0$)
2. Omitted variable correlated with included variable ($\text{Cov}(X_1, X_2) \neq 0$)

If either condition fails, no bias!
:::

**Example**: Returns to education

```{r}
#| label: ovb-example
# Short regression (omits ability)
short <- lm(log(wage) ~ educ, data = wage1)

# Long regression (includes IQ as proxy for ability)
long <- lm(log(wage) ~ educ + IQ, data = wage1)

modelsummary(
  list("Short" = short, "Long" = long),
  stars = TRUE
)
```

Education coefficient is **upward biased** in short regression (ability omitted).

### Irrelevant Variables {#sec-irrelevant}

Including **irrelevant** variables ($\beta_j = 0$):
- Estimates remain **unbiased**
- Standard errors **increase** (less efficient)
- R² may increase slightly

**Rule**: Include variables suggested by theory, even if insignificant.

### Specification Tests {#sec-spec-tests}

::: {.panel-tabset}

## RESET Test

Ramsey's RESET tests for omitted variables/functional form:

```{r}
#| label: reset-test
library(lmtest)

# RESET test (add powers of fitted values)
resettest(wage_model, power = 2:3)
```

## Davidson-MacKinnon Test

Compare two non-nested models:

```{r}
#| label: jtest
#| eval: false
# J-test for non-nested models
model1 <- lm(log(wage) ~ educ + exper + I(exper^2), data = wage1)
model2 <- lm(log(wage) ~ log(educ) + log(exper), data = wage1)

jtest(model1, model2)
```

:::

### Model Selection Criteria {#sec-info-criteria}

::: {.panel-tabset}

## Adjusted R²

$$
\bar{R}^2 = 1 - \frac{SSR/(n-k-1)}{SST/(n-1)} = 1 - (1-R^2)\frac{n-1}{n-k-1}
$$ {#eq-adj-r2}

Penalizes adding variables. Choose model with **highest** $\bar{R}^2$.

## AIC

$$
AIC = \log(SSR/n) + 2(k+1)/n
$$ {#eq-aic}

Choose model with **lowest** AIC.

## BIC

$$
BIC = \log(SSR/n) + (k+1)\log(n)/n
$$ {#eq-bic}

Stronger penalty than AIC. Choose model with **lowest** BIC.

:::

```{r}
#| label: model-selection
# Compare models
model1 <- lm(log(wage) ~ educ, data = wage1)
model2 <- lm(log(wage) ~ educ + exper, data = wage1)
model3 <- lm(log(wage) ~ educ + exper + I(exper^2), data = wage1)
model4 <- lm(log(wage) ~ educ + exper + I(exper^2) + tenure, data = wage1)

# Selection criteria
modelsummary(
  list("M1" = model1, "M2" = model2, "M3" = model3, "M4" = model4),
  gof_map = c("r.squared", "adj.r.squared", "AIC", "BIC", "nobs")
)
```

## Quantile Regression {#sec-quantile}

OLS estimates the **conditional mean**: $E(Y|X)$

**Quantile regression** estimates conditional **quantiles**: $Q_\tau(Y|X)$

$$
Q_\tau(Y|X) = X'\beta_\tau
$$

Where $\tau \in (0,1)$ is the quantile (e.g., $\tau = 0.5$ is the median).

::: {.callout-note}
## Why Quantile Regression?
- Robust to outliers
- See how $X$ affects **entire distribution** of $Y$
- Effect of $X$ may differ across quantiles (e.g., education matters more at top of wage distribution)
:::

```{r}
#| label: quantile-reg
library(quantreg)

# Quantile regressions at 10th, 50th, 90th percentiles
qr_10 <- rq(log(wage) ~ educ + exper + I(exper^2), tau = 0.10, data = wage1)
qr_50 <- rq(log(wage) ~ educ + exper + I(exper^2), tau = 0.50, data = wage1)
qr_90 <- rq(log(wage) ~ educ + exper + I(exper^2), tau = 0.90, data = wage1)

# OLS for comparison
ols <- lm(log(wage) ~ educ + exper + I(exper^2), data = wage1)

# Compare
modelsummary(
  list("Q10" = qr_10, "Q50 (Median)" = qr_50, "Q90" = qr_90, "OLS (Mean)" = ols),
  stars = TRUE,
  gof_map = "nobs"
)
```

**Interpretation**: Return to education is highest at the 90th percentile (`r round(100 * coef(qr_90)["educ"], 2)`%) compared to 10th percentile (`r round(100 * coef(qr_10)["educ"], 2)`%).

```{r}
#| label: fig-quantile
#| fig-cap: "Quantile Regression Coefficients for Education"
#| fig-width: 7
#| fig-height: 5
# Coefficient plot across quantiles
taus <- seq(0.1, 0.9, by = 0.05)
qr_fits <- lapply(taus, function(tau) {
  rq(log(wage) ~ educ + exper + I(exper^2), tau = tau, data = wage1)
})

coefs <- sapply(qr_fits, function(fit) coef(fit)["educ"])

plot_data <- data.frame(tau = taus, coef = coefs)

ggplot(plot_data, aes(x = tau, y = coef)) +
  geom_line(color = "darkblue", linewidth = 1) +
  geom_hline(yintercept = coef(ols)["educ"], linetype = "dashed", color = "red") +
  labs(
    x = "Quantile (τ)",
    y = "Education Coefficient",
    title = "Return to Education Across Wage Distribution",
    subtitle = "Dashed line = OLS estimate"
  ) +
  theme_minimal()
```

## Summary {#sec-lm-summary}

::: {.callout-note}
## Key Takeaways

**OLS Estimation**:
- Minimizes sum of squared residuals
- Under Gauss-Markov assumptions: BLUE
- Easy to compute and interpret

**Inference**:
- t-tests for individual coefficients
- F-tests for joint hypotheses
- Robust standard errors for heteroskedasticity

**Model Building**:
- Include variables suggested by theory
- Check for omitted variable bias
- Test functional form
- Use information criteria for selection

**Extensions**:
- Dummy variables for qualitative factors
- Interactions for differential effects
- Logs and polynomials for nonlinearity
- Quantile regression for distributional effects

**Common Issues**:
- Heteroskedasticity → use robust SEs
- Multicollinearity → larger SEs
- Omitted variables → bias
- Endogeneity → biased and inconsistent (need IV!)
:::

## Further Reading {#sec-lm-references}

- @wooldridge2015introductory - Chapters 2-9
- @heiss2016using - Practical R implementation
- @greene2003econometric - Advanced theory
- @angrist2009mostly - Practical econometric thinking

---

::: {.callout-tip}
## Next Chapter
See @sec-panel for **Panel Data Models** - exploiting repeated observations over time.
:::
