# Linear Regression {#sec-lm}

::: {.callout-note}
## Chapter Status
This is the largest chapter (2800+ lines). Full native Quarto conversion is in progress.

For complete content, the original LM.Rmd is included via child chunk below.
:::

```{r}
#| label: setup-lm
#| include: false
# Packages loaded in before-chapter.R
```

## Simple Linear Regression {#sec-slr}

The simple linear regression model:

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
$$ {#eq-slr}

Where:

- $Y_i$: Dependent variable
- $X_i$: Independent variable
- $\beta_0$: Intercept
- $\beta_1$: Slope coefficient
- $\epsilon_i$: Error term with $E(\epsilon_i|X_i) = 0$

::: {.callout-important}
## OLS Assumptions (SLR)
1. **Linearity**: True relationship is linear
2. **Random sampling**: $(X_i, Y_i)$ are i.i.d.
3. **No perfect collinearity**: Variation in $X$
4. **Zero conditional mean**: $E(\epsilon|X) = 0$
5. **Homoskedasticity**: $\text{Var}(\epsilon|X) = \sigma^2$
:::

### OLS Estimators {#sec-ols-slr}

$$
\hat{\beta}_1 = \frac{\sum(X_i - \bar{X})(Y_i - \bar{Y})}{\sum(X_i - \bar{X})^2} = \frac{\text{Cov}(X,Y)}{\text{Var}(X)}
$$ {#eq-beta1}

$$
\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1 \bar{X}
$$ {#eq-beta0}

## Multiple Linear Regression {#sec-mlr}

With $k$ explanatory variables:

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + \epsilon_i
$$ {#eq-mlr}

In matrix form:

$$
\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$ {#eq-mlr-matrix}

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}
$$ {#eq-ols-matrix}

::: {.callout-important}
## MLR Assumptions (Gauss-Markov)
1. **Linear in parameters**
2. **Random sampling**
3. **No perfect multicollinearity**
4. **Zero conditional mean**: $E(\epsilon|\mathbf{X}) = 0$
5. **Homoskedasticity**: $\text{Var}(\epsilon|\mathbf{X}) = \sigma^2$

Under these: OLS is **BLUE** (Best Linear Unbiased Estimator)
:::

## Inference {#sec-inference}

### Standard Errors {#sec-se}

$$
\text{SE}(\hat{\beta}_j) = \sqrt{\frac{\hat{\sigma}^2}{SST_j(1-R_j^2)}}
$$ {#eq-se}

### t-Tests {#sec-t-test}

Test $H_0: \beta_j = 0$:

$$
t = \frac{\hat{\beta}_j}{\text{SE}(\hat{\beta}_j)} \sim t_{n-k-1}
$$ {#eq-t-stat}

### F-Tests {#sec-f-test}

Test multiple restrictions $H_0: \beta_1 = \beta_2 = \cdots = \beta_k = 0$:

$$
F = \frac{(SSR_r - SSR_{ur})/q}{SSR_{ur}/(n-k-1)} \sim F_{q,n-k-1}
$$ {#eq-f-stat}

## Heteroskedasticity {#sec-hetero}

When $\text{Var}(\epsilon_i|X_i) = \sigma_i^2$ varies with $X$:

::: {.callout-warning}
## Consequences of Heteroskedasticity
- OLS still **unbiased** and **consistent**
- Standard errors are **biased**
- t-tests and F-tests **invalid**

**Solution**: Use robust (heteroskedasticity-consistent) standard errors
:::

### Robust Standard Errors {#sec-robust-se}

```{r}
#| label: robust-se-example
#| eval: false
library(lmtest)
library(sandwich)

# OLS model
model <- lm(y ~ x1 + x2, data = mydata)

# Robust SEs (HC1)
coeftest(model, vcov = vcovHC(model, type = "HC1"))
```

## Model Selection {#sec-model-selection}

::: {.panel-tabset}

## Adjusted R²

$$
\bar{R}^2 = 1 - \frac{SSR/(n-k-1)}{SST/(n-1)}
$$ {#eq-adj-r2}

Penalizes adding variables

## AIC

$$
AIC = \log(SSR/n) + 2(k+1)/n
$$ {#eq-aic}

## BIC

$$
BIC = \log(SSR/n) + (k+1)\log(n)/n
$$ {#eq-bic}

Stronger penalty for model complexity

:::

## Detailed Content

::: {.callout-note}
## Full Chapter Conversion
The complete Linear Regression chapter covers:

- Simple and multiple regression
- Dummy variables and interactions
- Functional forms (log, polynomial, splines)
- Outliers and influential observations
- Heteroskedasticity testing and correction
- Specification tests
- Quantile regression

Full native Quarto conversion is planned for Phase 3.
:::

```{r}
#| child: "LM.Rmd"
```
