# Panel Data Models {#sec-panel}

::: {.callout-note}
## Chapter Overview
This chapter covers panel data econometrics, including:

- Pooled cross-sections and panel data structure
- Fixed effects and random effects models
- Testing procedures for panel data
- Dynamic panel data models
:::

```{r}
#| label: setup-panel
#| include: false
# Packages are loaded in before-chapter.R
# Additional panel-specific packages are already available:
# plm, pglm, fixest, lfe, lme4, pmdplyr
```

## Pooled Cross-Sections

Pooled cross sections consist of random samples from the same population at different points in time.

### Changes to Return to Education and the Gender Wage Gap

The data set `cps78_85` includes two pooled cross-sections for the year 1978 and 1985. The dummy variable **y85** is equal to one for observations in 1985 and 0 for observations in 1978.

We estimate a model for:

$$
\log(w) = \beta_0 + \delta_{0}y85 + \beta_1 educ + \delta_{1} (y85 \times educ) +  \beta_{2} exper +  \beta_{3} \frac{exper^2}{100} + \beta_{4} union + \beta_{5}female + \delta_{2} (y85 \times female) + \epsilon
$$ {#eq-wage-model}

```{r}
#| label: load-cps-data
#| message: false
data("cps78_85", package = "wooldridge")
head(cps78_85)
```

```{r}
#| label: pooled-ols
cps_ols <- lm(lwage ~ y85 * (educ + female) + exper + I((exper^2)/100) + union,
              data = cps78_85)
summary(cps_ols)
```

::: {.panel-tabset}

## Summary Output

```{r}
#| echo: false
library(broom)
tidy(cps_ols) %>%
  knitr::kable(digits = 4)
```

## Stargazer Table

```{r}
stargazer::stargazer(cps_ols, type = "text")
```

:::

### Interpreting Coefficients

**Return to education:**

```{r}
#| label: return-education
df_cps_ols <- tidy(cps_ols)
delta1 <- df_cps_ols %>%
  dplyr::filter(term == "y85:educ") %>%
  dplyr::select(estimate) %>%
  pull()
```

::: {.callout-tip}
## Interpretation
The return to education is estimated to have increased by `r round(delta1, 4)` log points between 1978 and 1985.
:::

**Gender wage gap:**

```{r}
#| label: gender-gap
beta5 <- df_cps_ols %>%
  dplyr::filter(term == "female") %>%
  dplyr::select(estimate) %>%
  pull()

delta5 <- df_cps_ols %>%
  dplyr::filter(term == "y85:female") %>%
  dplyr::select(estimate) %>%
  pull()
```

The gender wage gap in 1978 was `r round(beta5, 4)` log points.

In 1985, the gap decreased to `r round(beta5 + delta5, 4)` log points, representing a narrowing of `r round(-delta5, 4)` log points.

## Organizing Panel Data {#sec-organizing-panel}

::: {.callout-important}
## Panel Data Structure
For panel data methods to work correctly, data must be systematically organized. Usually panel data comes in a **long** structure where each row corresponds to one combination of unit $i$ and time $t$.
:::

The **plm**, **fixest**, and **lfe** packages are widely used for panel data in R:

- **plm**: Comprehensive panel data package with many estimators
- **fixest**: Fast fixed effects estimation (recommended for large datasets)
- **lfe**: Linear models with multiple fixed effects
- **lme4**: Multilevel and hierarchical models
- **pmdplyr**: Panel data manipulation with dplyr syntax

::: {.panel-tabset}

## plm Package

For unbalanced panels, specify a panel data frame with `pdata.frame(mydf, index = c("ivar", "tvar"))`. For balanced panels, you can simply define the index `pdata.frame(mydf, index = n)`.

```{r}
#| label: plm-organize
data("crime2", package = "wooldridge")
head(crime2, 5)
```

::: {.callout-note}
Wooldridge remarks that the study observations result in a balanced panel of 46 cities, properly sorted.
:::

```{r}
#| label: create-pdata
crime2.p <- plm::pdata.frame(crime2, index = 46)
```

```{r}
#| label: show-pdata
crime2.p[1:6, c("id", "time", "year", "pop", "crimes", "crmrte", "unem")]
```

## pmdplyr Package

The **pmdplyr** package extends dplyr for panel data:

```{r}
#| label: pmdplyr-organize
library(pmdplyr)
crime2 <- crime2 %>%
  as_pibble(.i = id, .t = year)

crime2
```

## fixest Package

**fixest** is the fastest option for large datasets:

```{r}
#| label: fixest-organize
#| eval: false
# fixest works directly with regular data frames
# No special panel data class needed
# Just specify panel structure in the formula
```

:::

### Panel-Specific Computations

```{r}
#| label: panel-dimensions
plm::pdim(crime2.p)
```

## Estimation Methods {#sec-estimation}

::: {.panel-tabset}

## Pooled OLS

Standard OLS ignoring panel structure:

```{r}
#| label: pooled-model
pooled <- plm::plm(log(crmrte) ~ d83 + d84 + d85 + d86 + d87 +
                  lprbarr + lprbconv + lprbpris + lavgsen + lpolpc,
                data = crime2.p, model = "pooling")
summary(pooled)
```

## Fixed Effects

Removes time-invariant unobserved heterogeneity:

```{r}
#| label: fe-model
fe_plm <- plm::plm(log(crmrte) ~ d83 + d84 + d85 + d86 + d87 +
                lprbarr + lprbconv + lprbpris + lavgsen + lpolpc,
              data = crime2.p, model = "within")
summary(fe_plm)
```

## Random Effects

Assumes individual effects uncorrelated with regressors:

```{r}
#| label: re-model
re_model <- plm::plm(log(crmrte) ~ d83 + d84 + d85 + d86 + d87 +
                  lprbarr + lprbconv + lprbpris + lavgsen + lpolpc,
                data = crime2.p, model = "random")
summary(re_model)
```

## First Differences

Eliminates individual effects by differencing:

```{r}
#| label: fd-model
fd_model <- plm::plm(log(crmrte) ~ d83 + d84 + d85 + d86 + d87 +
                  lprbarr + lprbconv + lprbpris + lavgsen + lpolpc,
                data = crime2.p, model = "fd")
summary(fd_model)
```

:::

::: {.callout-tip}
## Which Estimator to Use?
- **Pooled OLS**: Only if no individual effects
- **Fixed Effects**: Consistent even if effects correlated with X
- **Random Effects**: More efficient if effects uncorrelated with X
- **First Differences**: Useful when strict exogeneity fails
:::

## Testing {#sec-panel-tests}

::: {.panel-tabset}

## F Test

Test if fixed effects are needed:

```{r}
#| label: f-test
plm::pFtest(fe_plm, pooled)
```

## LM Test

Breusch-Pagan test for random effects:

```{r}
#| label: lm-test
plm::plmtest(pooled, type = "bp")
```

## Hausman Test

Choose between FE and RE:

```{r}
#| label: hausman
plm::phtest(fe_plm, re_model)
```

:::

::: {.callout-important}
## Hausman Test Interpretation
- **p < 0.05**: Use Fixed Effects (reject RE)
- **p ≥ 0.05**: Use Random Effects (more efficient)
:::

## Robust Standard Errors {#sec-robust}

Panel data requires cluster-robust standard errors:

```{r}
#| label: robust-se
lmtest::coeftest(fe_plm, vcov = plm::vcovHC(fe_plm, type = "HC1", cluster = "group"))
```

::: {.callout-warning}
Always use cluster-robust standard errors with panel data!
:::

## Dynamic Panel Models {#sec-dynamic}

Include lagged dependent variables:

```{r}
#| label: dynamic-panel
#| eval: false
# Arellano-Bond GMM estimator
library(plm)
pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
       lag(log(capital), 0:1) | lag(log(emp), 2:99),
     data = EmplUK, effect = "twoways", model = "onestep")
```

::: {.callout-note}
## Dynamic Panel Estimators
- **Arellano-Bond**: First-difference GMM
- **Blundell-Bond**: System GMM (more efficient)
:::

## Summary

Key takeaways from panel data analysis:

1. Check panel structure with `pdim()` and `pvar()`
2. Use Hausman test to choose between FE and RE
3. Always use cluster-robust standard errors
4. For dynamic panels, use GMM estimators
5. **fixest** package is fastest for large datasets

## Further Reading

- Wooldridge (2010): *Econometric Analysis of Cross Section and Panel Data*
- [plm vignette](https://cran.r-project.org/web/packages/plm/vignettes/plmPackage.html)
- [fixest documentation](https://lrberge.github.io/fixest/)
