# Special Topics {#sec-special}

::: {.callout-note}
## Chapter Overview
This chapter covers specialized econometric methods:

- Difference-in-Differences estimation
- Regression Discontinuity Design
- Instrumental Variables applications
- Synthetic Control Methods
:::

```{r}
#| label: setup-special
#| include: false
# Standard packages loaded in before-chapter.R
```

## Difference-in-Differences {#sec-did}

Difference-in-Differences (DiD) is an important quasi-experimental method for identifying causal effects. It estimates the effect of a policy intervention by comparing the change over time in outcomes between treated and control groups.

::: {.callout-important}
## DiD Assumptions
1. **Parallel Trends**: Treatment and control groups would have followed parallel trends absent treatment
2. **No anticipation**: Units don't change behavior before treatment
3. **SUTVA**: Treatment of one unit doesn't affect others
:::

### Basic DiD Framework

The DiD estimator compares the difference in outcomes before and after treatment between treatment and control groups:

$$
\text{DiD} = [E(Y_{1,treated}) - E(Y_{0,treated})] - [E(Y_{1,control}) - E(Y_{0,control})]
$$ {#eq-did-estimator}

### Regression Specification

$$
Y_{it} = \beta_0 + \beta_1 \text{Treat}_i + \beta_2 \text{Post}_t + \beta_3 (\text{Treat}_i \times \text{Post}_t) + \epsilon_{it}
$$ {#eq-did-regression}

Where $\beta_3$ is the DiD estimate of the treatment effect.

### Example: Garbage Incinerator and Housing Prices

We examine whether construction of a garbage incinerator affected nearby house values.

```{r}
#| label: load-kielmc
data("kielmc", package = "wooldridge")
head(kielmc)
```

::: {.panel-tabset}

## Separate Regressions

Estimate effects for each year separately:

```{r}
#| label: separate-regressions
# 1978 (before incinerator)
model_1978 <- lm(rprice ~ nearinc, data = kielmc, subset = (year == 1978))

# 1981 (after incinerator)
model_1981 <- lm(rprice ~ nearinc, data = kielmc, subset = (year == 1981))

# Compare coefficients
data.frame(
  Year = c(1978, 1981),
  Coefficient = c(coef(model_1978)["nearinc"],
                  coef(model_1981)["nearinc"])
)
```

## DiD Regression

Joint regression with interaction:

```{r}
#| label: did-model
did_model <- lm(rprice ~ nearinc * y81, data = kielmc)
lmtest::coeftest(did_model)
```

::: {.callout-tip}
## Interpretation
The coefficient on `nearinc:y81` is the DiD estimate: the differential effect of being near the incinerator after it was built.
:::

## One-Tailed Test

Test if incinerator decreased house values (H₁: β < 0):

```{r}
#| label: one-tailed-test
summary_did <- summary(did_model)
pt(coef(summary_did)[, 3], did_model$df, lower = TRUE)
```

:::

### Log Specification with Controls

A logarithmic specification implies constant percentage effects:

```{r}
#| label: did-log-models
# Log specification without controls
did_log1 <- lm(log(rprice) ~ nearinc * y81, data = kielmc)

# Log specification with controls
did_log2 <- lm(log(rprice) ~ nearinc * y81 + age + I(age^2) +
                 log(intst) + log(land) + log(area) +
                 rooms + baths,
               data = kielmc)
```

```{r}
#| label: tbl-did-comparison
#| tbl-cap: "Difference-in-Differences estimates"
huxtable::huxreg(did_log1, did_log2,
                 statistics = c("N" = "nobs", "R²" = "r.squared"))
```

::: {.callout-note}
## DiD Results Interpretation
With log(rprice) as outcome:

- Coefficient ≈ percentage change in house value
- Controls reduce omitted variable bias
- Standard errors should be clustered if repeated observations
:::

## Recent Developments in DiD {#sec-modern-did}

::: {.callout-warning}
## Staggered Adoption & Two-Way Fixed Effects
Recent research shows traditional two-way fixed effects DiD can be biased with:

- **Staggered treatment** adoption
- **Heterogeneous** treatment effects
- **Dynamic** effects

**Solutions**:
- Callaway & Sant'Anna (2021) estimator
- Sun & Abraham (2021) interaction-weighted estimator
- Borusyak et al. (2022) imputation estimator
:::

### Modern DiD Packages

```{r}
#| label: modern-did
#| eval: false
# Callaway & Sant'Anna
library(did)
out <- att_gt(yname = "y",
              tname = "year",
              idname = "id",
              gname = "first_treat",
              data = mydata)

# Sun & Abraham
library(fixest)
feols(y ~ x | id + year, data = mydata,
      sunab = TRUE)
```

## Other Special Topics {#sec-other-topics}

### Regression Discontinuity Design (RDD)

RDD exploits discontinuities in treatment assignment:

```{r}
#| label: rdd-example
#| eval: false
library(rdd)
# Sharp RD
rd_est <- RDestimate(y ~ running_var, data = mydata, cutpoint = 0)
```

::: {.callout-note}
**RDD Types**:
- **Sharp RD**: Treatment changes discontinuously at cutoff
- **Fuzzy RD**: Treatment probability changes at cutoff
:::

### Synthetic Control Methods

For comparative case studies with one treated unit:

```{r}
#| label: synth-example
#| eval: false
library(Synth)
# Create synthetic control
synth_out <- synth(
  data.prep.obj = dataprep_out,
  method = "BFGS"
)
```

## Summary

Special topics covered:

✅ **Difference-in-Differences**: Parallel trends assumption critical
✅ **Modern DiD**: Account for staggered adoption and heterogeneity
✅ **RDD**: Local treatment effects at discontinuity
✅ **Synthetic Control**: Comparative case studies

::: {.callout-tip}
## Causal Inference Best Practices
1. **Clearly state identifying assumptions**
2. **Test assumptions when possible** (e.g., parallel trends)
3. **Conduct robustness checks**
4. **Report standard errors** (clustered appropriately)
5. **Visualize treatment effects**
:::

## Further Reading

- Angrist & Pischke (2009): *Mostly Harmless Econometrics*
- Cunningham (2021): [*Causal Inference: The Mixtape*](https://mixtape.scunning.com/)
- Huntington-Klein (2021): [*The Effect*](https://theeffectbook.net/)
- [did package vignette](https://bcallaway11.github.io/did/)
