# Time Series Analysis {#sec-timeseries}

```{r}
#| label: setup-ts
#| include: false
# Packages loaded in before-chapter.R
```

::: {.callout-note}
## Chapter Overview
This chapter covers time series econometrics:

- Time series objects in R (ts, zoo, xts)
- Stationarity and unit root testing
- ARIMA modeling and forecasting
- Seasonal decomposition
- Finite Distributed Lag (FDL) models
- Vector Autoregression (VAR) models
:::

## Introduction to Time Series {#sec-ts-intro}

**Time series data**: Observations indexed by time $t = 1, 2, ..., T$

$$
\{Y_t\}_{t=1}^T = Y_1, Y_2, ..., Y_T
$$

::: {.callout-tip}
## Time Series vs Cross-Sectional Data

**Time series**: Same unit observed over time
- Example: GDP of Estonia from 1995-2023
- Key issue: **Serial correlation** (observations not independent)
- Order matters!

**Cross-sectional**: Different units at one time
- Example: Wages of 1000 workers in 2023
- Assumes independence across units
- Order doesn't matter
:::

### Time Series Objects in R {#sec-ts-objects}

::: {.panel-tabset}

## ts Object

Regular time series (equal spacing):

```{r}
#| label: ts-object-demo
# Create monthly time series starting Jan 1996
example_ts <- ts(
  data = rnorm(48),  # 4 years of monthly data
  start = c(1996, 1),
  frequency = 12     # 12 months per year
)

# Properties
str(example_ts)
frequency(example_ts)
cycle(example_ts)[1:12]  # Which month?
time(example_ts)[1:12]   # Decimal year
```

## zoo Object

Irregular time series (flexible dates):

```{r}
#| label: zoo-object-demo
library(zoo)

# Irregular dates
dates <- as.Date(c("2020-01-01", "2020-01-05", "2020-01-12", "2020-02-03"))
values <- c(100, 102, 98, 105)

z <- zoo(values, dates)
z

# Can handle gaps!
```

## xts Object

Extended time series (most powerful):

```{r}
#| label: xts-object-demo
library(xts)

# xts = zoo + extra features
x <- xts(values, dates)
x

# Subsetting by date
x["2020-01"]  # All January 2020
```

:::

**Which to use?**
- **ts**: Regular data (monthly, quarterly, annual)
- **zoo**: Irregular data with dates
- **xts**: Advanced time series operations

### Example: Estonian CPI {#sec-cpi-example}

```{r}
#| label: load-cpi
#| eval: false
library(eurostat)

# Download monthly CPI from Eurostat
cpi <- get_eurostat(
  "prc_hicp_midx",
  filters = list(geo = "EE", coicop = "CP00", unit = "I15")
) %>%
  filter(time >= as.Date("1996-01-01")) %>%
  select(time, values) %>%
  rename(cpi = values, Date = time)

# Convert to ts object
cpi_ts <- ts(cpi$cpi, start = c(1996, 1), frequency = 12)
```

```{r}
#| label: fig-cpi
#| fig-cap: "Estonian Consumer Price Index"
#| eval: false
#| fig-width: 8
#| fig-height: 5
plot(cpi_ts,
  main = "Estonian CPI (1996-present)",
  ylab = "Index (2015=100)",
  xlab = "Year"
)
```

## Stationarity {#sec-stationarity}

### Definition {#sec-stationarity-def}

A time series $\{Y_t\}$ is **weakly stationary** (or covariance stationary) if:

1. **Constant mean**: $E(Y_t) = \mu$ for all $t$
2. **Constant variance**: $\text{Var}(Y_t) = \sigma^2$ for all $t$
3. **Autocovariance depends only on lag**: $\text{Cov}(Y_t, Y_{t-h}) = \gamma_h$ for all $t$

::: {.callout-important}
## Why Does Stationarity Matter?

**Non-stationary data**:
- Standard regression assumptions violated
- Spurious regression problem
- Invalid inference

**Solution**: Transform to stationarity
- Take differences: $\Delta Y_t = Y_t - Y_{t-1}$
- Take logs: $\log(Y_t)$
- Detrend: Remove deterministic trend
:::

### Examples {#sec-stationarity-examples}

::: {.panel-tabset}

## Stationary: White Noise

```{r}
#| label: fig-white-noise
#| fig-cap: "White Noise (Stationary)"
#| fig-width: 7
#| fig-height: 4
set.seed(123)
wn <- ts(rnorm(200), frequency = 1)

par(mfrow = c(1, 2))
plot(wn, main = "White Noise", ylab = "Y_t")
acf(wn, main = "ACF")
```

**Properties**:
- Mean = 0 (constant)
- Variance constant
- No autocorrelation

## Non-Stationary: Random Walk

```{r}
#| label: fig-random-walk
#| fig-cap: "Random Walk (Non-Stationary)"
#| fig-width: 7
#| fig-height: 4
set.seed(123)
rw <- ts(cumsum(rnorm(200)), frequency = 1)

par(mfrow = c(1, 2))
plot(rw, main = "Random Walk", ylab = "Y_t")
acf(rw, main = "ACF (slowly decaying)")
```

**Properties**:
- $Y_t = Y_{t-1} + \epsilon_t$
- Variance increases over time
- High persistence (ACF decays slowly)

## Non-Stationary: Trend

```{r}
#| label: fig-trend
#| fig-cap: "Deterministic Trend (Non-Stationary)"
#| fig-width: 7
#| fig-height: 4
t <- 1:200
trend <- ts(10 + 0.5 * t + rnorm(200, sd = 5), frequency = 1)

par(mfrow = c(1, 2))
plot(trend, main = "Linear Trend", ylab = "Y_t")
acf(trend, main = "ACF")
```

**Properties**:
- $Y_t = \alpha + \beta t + \epsilon_t$
- Mean increases over time
- Solution: Detrend or difference

:::

## Unit Root Testing {#sec-unit-root}

### The Unit Root Problem {#sec-unit-root-problem}

Consider the AR(1) model:

$$
Y_t = \rho Y_{t-1} + \epsilon_t
$$

- If $|\rho| < 1$: **Stationary** (mean-reverting)
- If $\rho = 1$: **Unit root** (random walk, non-stationary)
- If $|\rho| > 1$: **Explosive** (rare in economics)

**Testing**: $H_0: \rho = 1$ (unit root) vs $H_1: |\rho| < 1$ (stationary)

::: {.callout-warning}
## Spurious Regression
If $Y_t$ and $X_t$ both have unit roots but are **unrelated**, OLS regression of $Y_t$ on $X_t$ will show:
- High R² (often > 0.7)
- Significant t-statistics
- But the relationship is **spurious**!

**Solution**: Test for unit roots first, then difference if needed
:::

### Unit Root Tests {#sec-unit-root-tests}

::: {.panel-tabset}

## Augmented Dickey-Fuller (ADF)

Tests $H_0$: Unit root (non-stationary)

$$
\Delta Y_t = \alpha + \beta t + \gamma Y_{t-1} + \sum_{i=1}^p \delta_i \Delta Y_{t-i} + \epsilon_t
$$

Test $H_0: \gamma = 0$ vs $H_1: \gamma < 0$

```{r}
#| label: adf-test-example
library(urca)

# Simulate random walk (has unit root)
set.seed(123)
y_rw <- cumsum(rnorm(100))

# ADF test
adf_test <- ur.df(y_rw, type = "drift", lags = 4)
summary(adf_test)
```

**Interpretation**:
- Test statistic < critical value → Reject $H_0$ (stationary)
- Test statistic > critical value → Cannot reject $H_0$ (unit root)

## Phillips-Perron (PP)

Non-parametric alternative to ADF:

```{r}
#| label: pp-test
library(tseries)

# PP test
pp.test(y_rw)
```

**Advantages**:
- Robust to heteroskedasticity
- No need to specify lag length

## KPSS Test

**Reverses null hypothesis**: $H_0$: Stationary vs $H_1$: Unit root

```{r}
#| label: kpss-test
# KPSS test
kpss.test(y_rw)
```

**Strategy**: Use both ADF and KPSS
- ADF fails to reject & KPSS rejects → unit root
- ADF rejects & KPSS fails to reject → stationary
- Both reject or both fail to reject → ambiguous (use more lags/tests)

:::

### First Differences {#sec-differences}

If $Y_t$ has a unit root, $\Delta Y_t = Y_t - Y_{t-1}$ is often stationary:

```{r}
#| label: differencing
# Random walk (non-stationary)
y_rw <- cumsum(rnorm(100))

# First difference (stationary)
dy_rw <- diff(y_rw)

# Test differenced series
adf.test(dy_rw)
```

::: {.callout-note}
## I(d) Notation
- **I(0)**: Integrated of order 0 = Stationary
- **I(1)**: Integrated of order 1 = Needs 1 difference to be stationary
- **I(2)**: Needs 2 differences (rare)

Most economic time series are **I(1)**: GDP, prices, wages, stock prices
:::

## ARIMA Models {#sec-arima}

### ARIMA Components {#sec-arima-components}

**ARIMA(p,d,q)**: AutoRegressive Integrated Moving Average

$$
\phi(L)(1-L)^d Y_t = \theta(L) \epsilon_t
$$

Where:
- **AR(p)**: $Y_t = \phi_1 Y_{t-1} + \cdots + \phi_p Y_{t-p} + \epsilon_t$
- **I(d)**: Differencing $d$ times
- **MA(q)**: $\epsilon_t = \theta_1 \epsilon_{t-1} + \cdots + \theta_q \epsilon_{t-q}$

::: {.panel-tabset}

## AR(p) Models

**Autoregressive**: $Y_t$ depends on its own past values

$$
Y_t = c + \phi_1 Y_{t-1} + \phi_2 Y_{t-2} + \cdots + \phi_p Y_{t-p} + \epsilon_t
$$

**AR(1) example**:
```{r}
#| label: ar1-example
# Simulate AR(1): Y_t = 0.7*Y_{t-1} + ε_t
set.seed(123)
ar1 <- arima.sim(model = list(ar = 0.7), n = 200)

par(mfrow = c(1, 2))
plot(ar1, main = "AR(1) with φ=0.7")
acf(ar1, main = "ACF (exponential decay)")
```

**Identifying AR(p)**:
- **ACF**: Decays gradually
- **PACF**: Cuts off after lag p

## MA(q) Models

**Moving Average**: $Y_t$ depends on past errors

$$
Y_t = \mu + \epsilon_t + \theta_1 \epsilon_{t-1} + \theta_2 \epsilon_{t-2} + \cdots + \theta_q \epsilon_{t-q}
$$

**MA(1) example**:
```{r}
#| label: ma1-example
# Simulate MA(1): Y_t = ε_t + 0.5*ε_{t-1}
ma1 <- arima.sim(model = list(ma = 0.5), n = 200)

par(mfrow = c(1, 2))
plot(ma1, main = "MA(1) with θ=0.5")
acf(ma1, main = "ACF (cuts off at lag 1)")
```

**Identifying MA(q)**:
- **ACF**: Cuts off after lag q
- **PACF**: Decays gradually

## ARMA(p,q) Models

Combination of AR and MA:

$$
Y_t = c + \phi_1 Y_{t-1} + \cdots + \phi_p Y_{t-p} + \epsilon_t + \theta_1 \epsilon_{t-1} + \cdots + \theta_q \epsilon_{t-q}
$$

```{r}
#| label: arma-example
# ARMA(2,1)
arma <- arima.sim(model = list(ar = c(0.5, 0.3), ma = 0.4), n = 200)

plot(arma, main = "ARMA(2,1)")
```

:::

### Model Identification {#sec-arima-id}

**Box-Jenkins methodology**:

1. **Check stationarity**: Unit root tests
2. **Difference if needed**: Make stationary
3. **Examine ACF/PACF**: Identify p, q
4. **Estimate model**: Use `arima()` or `auto.arima()`
5. **Diagnostic checking**: Residual tests
6. **Forecasting**

::: {.callout-tip}
## ACF and PACF Patterns

| Model  | ACF                    | PACF                   |
|--------|------------------------|------------------------|
| AR(p)  | Decays exponentially   | Cuts off after lag p   |
| MA(q)  | Cuts off after lag q   | Decays exponentially   |
| ARMA   | Decays gradually       | Decays gradually       |
:::

### Example: ARIMA Estimation {#sec-arima-estimation}

```{r}
#| label: arima-estimation-example
#| eval: false
library(forecast)

# Load data (e.g., quarterly GDP growth)
gdp_growth <- diff(log(gdp)) * 100  # Percentage growth rate

# Check stationarity
adf.test(gdp_growth)

# ACF and PACF
par(mfrow = c(1, 2))
acf(gdp_growth)
pacf(gdp_growth)

# Fit ARIMA model
fit <- arima(gdp_growth, order = c(2, 0, 1))
summary(fit)

# Or use auto.arima (automatic selection)
auto_fit <- auto.arima(gdp_growth, seasonal = FALSE)
summary(auto_fit)

# Diagnostic checking
checkresiduals(fit)

# Forecast
forecast_values <- forecast(fit, h = 8)  # 8 quarters ahead
plot(forecast_values)
```

### Model Selection {#sec-arima-selection}

::: {.panel-tabset}

## AIC / BIC

**Akaike Information Criterion**:
$$AIC = -2\log(L) + 2k$$

**Bayesian Information Criterion**:
$$BIC = -2\log(L) + k\log(T)$$

Where $L$ = likelihood, $k$ = # parameters, $T$ = sample size

```{r}
#| label: model-selection
#| eval: false
# Compare models
model1 <- arima(y, order = c(1, 0, 0))
model2 <- arima(y, order = c(2, 0, 0))
model3 <- arima(y, order = c(1, 0, 1))

AIC(model1, model2, model3)
BIC(model1, model2, model3)

# Choose lowest AIC/BIC
```

## auto.arima()

Automatic model selection:

```{r}
#| label: auto-arima
#| eval: false
# Searches over different (p,d,q) combinations
best_model <- auto.arima(
  y,
  max.p = 5,
  max.q = 5,
  max.d = 2,
  seasonal = FALSE,
  ic = "aicc"  # AICc (corrected AIC for small samples)
)

summary(best_model)
```

:::

## Seasonal Decomposition {#sec-seasonal}

### Seasonal Patterns {#sec-seasonal-patterns}

Many time series have **seasonal patterns**: regular fluctuations within a year

**Examples**:
- Retail sales (high in December)
- Ice cream sales (high in summer)
- Flu cases (high in winter)

**Decomposition**:
$$
Y_t = T_t + S_t + R_t
$$

Where:
- $T_t$: Trend component
- $S_t$: Seasonal component
- $R_t$: Remainder (irregular)

### STL Decomposition {#sec-stl}

**STL**: Seasonal and Trend decomposition using Loess

```{r}
#| label: stl-example
#| eval: false
# Load monthly data
data(AirPassengers)

# STL decomposition
stl_decomp <- stl(AirPassengers, s.window = "periodic")

# Plot components
plot(stl_decomp)

# Extract components
trend <- stl_decomp$time.series[, "trend"]
seasonal <- stl_decomp$time.series[, "seasonal"]
remainder <- stl_decomp$time.series[, "remainder"]
```

### Seasonal ARIMA {#sec-sarima}

**SARIMA(p,d,q)(P,D,Q)s**: Seasonal ARIMA

$$
\phi(L)\Phi(L^s)(1-L)^d(1-L^s)^D Y_t = \theta(L)\Theta(L^s)\epsilon_t
$$

```{r}
#| label: sarima-example
#| eval: false
# Seasonal ARIMA for monthly data
sarima_fit <- arima(
  y,
  order = c(1, 1, 1),              # Non-seasonal (p,d,q)
  seasonal = list(order = c(1, 1, 1), period = 12)  # Seasonal (P,D,Q)_s
)

# Or use auto.arima
auto_sarima <- auto.arima(y, seasonal = TRUE)
```

## Finite Distributed Lag Models {#sec-fdl}

### FDL Model {#sec-fdl-model}

**Distributed lag**: Current $Y$ depends on current and past values of $X$

$$
Y_t = \alpha + \beta_0 X_t + \beta_1 X_{t-1} + \cdots + \beta_q X_{t-q} + \epsilon_t
$$

**Example**: Effect of advertising on sales
- Today's sales affected by today's ad spending
- But also last week's, last month's, etc.

::: {.callout-note}
## Impact Multipliers
- **Impact multiplier**: $\beta_0$ (immediate effect)
- **Lag-k multiplier**: $\beta_k$ (effect after k periods)
- **Cumulative multiplier**: $\sum_{i=0}^q \beta_i$ (total long-run effect)
- **Mean lag**: $\frac{\sum_{i=0}^q i\beta_i}{\sum_{i=0}^q \beta_i}$ (average delay)
:::

### Example: FDL Estimation {#sec-fdl-example}

```{r}
#| label: fdl-estimation
#| eval: false
library(dynlm)

# Create lagged variables
# Y_t = α + β0*X_t + β1*X_{t-1} + β2*X_{t-2} + ε_t

fdl_model <- dynlm(Y ~ L(X, 0) + L(X, 1) + L(X, 2), data = ts_data)
summary(fdl_model)

# Or more compactly
fdl_model <- dynlm(Y ~ L(X, 0:2), data = ts_data)

# Test if lags are jointly significant
library(car)
linearHypothesis(fdl_model, c("L(X, 1) = 0", "L(X, 2) = 0"))
```

### Choosing Lag Length {#sec-lag-selection}

::: {.panel-tabset}

## Information Criteria

```{r}
#| label: lag-selection-ic
#| eval: false
# Try different lag lengths
lag_0 <- dynlm(Y ~ L(X, 0), data = ts_data)
lag_1 <- dynlm(Y ~ L(X, 0:1), data = ts_data)
lag_2 <- dynlm(Y ~ L(X, 0:2), data = ts_data)
lag_3 <- dynlm(Y ~ L(X, 0:3), data = ts_data)

# Compare AIC
AIC(lag_0, lag_1, lag_2, lag_3)
```

## F-Test

```{r}
#| label: lag-selection-f
#| eval: false
# Test if additional lag is significant
restricted <- dynlm(Y ~ L(X, 0:2), data = ts_data)
unrestricted <- dynlm(Y ~ L(X, 0:3), data = ts_data)

anova(restricted, unrestricted)
```

:::

## Vector Autoregression (VAR) {#sec-var}

### VAR Model {#sec-var-model}

**VAR**: Multiple time series, each depending on lags of all series

$$
\begin{bmatrix} Y_{1t} \\ Y_{2t} \end{bmatrix} =
\begin{bmatrix} c_1 \\ c_2 \end{bmatrix} +
\begin{bmatrix} \phi_{11} & \phi_{12} \\ \phi_{21} & \phi_{22} \end{bmatrix}
\begin{bmatrix} Y_{1,t-1} \\ Y_{2,t-1} \end{bmatrix} +
\begin{bmatrix} \epsilon_{1t} \\ \epsilon_{2t} \end{bmatrix}
$$

**Example**: GDP and unemployment
- GDP affects unemployment
- Unemployment affects GDP
- Both depend on their own past values

### VAR Estimation {#sec-var-estimation}

```{r}
#| label: var-example
#| eval: false
library(vars)

# Combine variables into a ts object
var_data <- cbind(gdp, unemp, inflation)

# Select lag length
VARselect(var_data, lag.max = 8, type = "both")

# Estimate VAR(2)
var_model <- VAR(var_data, p = 2, type = "both")
summary(var_model)

# Granger causality test
causality(var_model, cause = "gdp")

# Impulse response functions
irf_gdp <- irf(var_model, impulse = "gdp", response = "unemp", n.ahead = 20)
plot(irf_gdp)

# Forecast
var_forecast <- predict(var_model, n.ahead = 8)
plot(var_forecast)
```

### Granger Causality {#sec-granger}

**Question**: Does $X$ help forecast $Y$?

**Granger causality test**: Does adding lags of $X$ improve forecast of $Y$?

::: {.callout-note}
## "Granger Causality" ≠ Causality
"X Granger-causes Y" means:
- X helps **predict** Y
- Does **NOT** mean X **causes** Y in causal sense
- Could be reverse causality, confounders, etc.
:::

```{r}
#| label: granger-test
#| eval: false
library(lmtest)

# Does X Granger-cause Y?
grangertest(Y ~ X, order = 4, data = ts_data)

# Interpretation:
# Small p-value → X Granger-causes Y (X helps predict Y)
# Large p-value → X does not Granger-cause Y
```

## Forecasting {#sec-forecasting}

### Forecast Evaluation {#sec-forecast-eval}

**Out-of-sample forecasting**:
1. Split data: Training set + Test set
2. Estimate model on training set
3. Forecast test set
4. Compare forecast to actual values

**Forecast errors**: $e_t = Y_t - \hat{Y}_t$

::: {.panel-tabset}

## Accuracy Measures

**Mean Absolute Error**:
$$MAE = \frac{1}{h}\sum_{t=1}^h |e_t|$$

**Root Mean Squared Error**:
$$RMSE = \sqrt{\frac{1}{h}\sum_{t=1}^h e_t^2}$$

**Mean Absolute Percentage Error**:
$$MAPE = \frac{100}{h}\sum_{t=1}^h \left|\frac{e_t}{Y_t}\right|$$

```{r}
#| label: forecast-accuracy
#| eval: false
library(forecast)

# Calculate accuracy
accuracy(forecast_obj, test_data)
```

## Forecast Intervals

95% prediction interval:
$$\hat{Y}_{T+h} \pm 1.96 \times SE(\hat{Y}_{T+h})$$

```{r}
#| label: forecast-intervals
#| eval: false
# Forecast with intervals
fc <- forecast(model, h = 12, level = c(80, 95))

plot(fc)
```

:::

### Exponential Smoothing {#sec-exp-smooth}

**Simple exponential smoothing**:
$$\hat{Y}_{T+1} = \alpha Y_T + (1-\alpha)\hat{Y}_T$$

Where $0 < \alpha < 1$ (smoothing parameter)

```{r}
#| label: exp-smooth
#| eval: false
library(forecast)

# Simple exponential smoothing
ses_model <- ses(y, h = 10)
plot(ses_model)

# Holt's linear trend
holt_model <- holt(y, h = 10)

# Holt-Winters seasonal
hw_model <- hw(y, seasonal = "additive", h = 12)
```

## Summary {#sec-ts-summary}

::: {.callout-note}
## Key Takeaways

**Stationarity**:
- Check for unit roots (ADF, PP, KPSS tests)
- Difference non-stationary series
- Most economic variables are I(1)

**ARIMA Models**:
- Identify (p,d,q) using ACF/PACF
- Estimate with `arima()` or `auto.arima()`
- Check residuals (white noise?)
- Forecast future values

**Seasonal Patterns**:
- Decompose: Trend + Seasonal + Remainder
- Use SARIMA for seasonal data
- Remove seasonality before modeling

**Distributed Lags**:
- Current Y depends on past X values
- Estimate with `dynlm`
- Interpret impact/cumulative multipliers

**VAR Models**:
- Multiple interrelated time series
- Granger causality tests
- Impulse response functions
- Multi-step forecasting

**Best Practices**:
- Always plot your data first!
- Test for stationarity
- Use information criteria (AIC/BIC)
- Validate with out-of-sample forecasts
- Report forecast intervals
:::

## Further Reading {#sec-ts-references}

- @wooldridge2015introductory - Chapters 10-12, 18
- @hamilton1994time - Classic time series reference
- @hyndman2018forecasting - Modern forecasting (free online!)
- @enders2014applied - Applied time series
- @shumway2017time - R-based time series

---

::: {.callout-tip}
## Next Chapter
See @sec-panel for **Panel Data Models** combining cross-sectional and time dimensions.
:::
